---
title: "phenology_MODIS_MCD12Q2"
author: "Salome Hayler"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, load libraries}
# library geodata -> access to shuttle radar topography mission (SRTM)
library(geodata)
library(MODISTools)
```

```{r}
# download SRTM data
# This stores file srtm_38_03.tif in 
# subfolder elevation of tempdir()
geodata::elevation_3s(
  lat = 46.6756, ## hier coordinaten zu boston ändern
  lon = 7.85480,
  path = tempdir()
)
```

```{r}
# read the downloaded data
# use file.path() to combine
# a directory path with a filename
dem <- terra::rast(
  file.path(
    tempdir(),
    "elevation",
    "srtm_38_03.tif"
  )
)
```

download MODIS land surface phenology product (MCD12Q2) 
```{r}
# download and save phenology data
phenology <- MODISTools::mt_subset(
  product = "MCD12Q2",
  lat = 46.6756,
  lon = 7.85480,
  band = "Greenup.Num_Modes_01",
  start = "2012-01-01",
  end = "2012-12-31",
  km_lr = 100,
  km_ab = 100,
  site_name = "swiss",
  internal = TRUE,
  progress = FALSE
)
```

preprocessing is needed because of: 
MODIS data comes as a tidy data frame
MODIS data might have missing values
DEM data extent is larger than MODIS coverage
Two non-matching grids (DEM ~ MODIS)

```{r}
#### preprocessing MODIS ####
# screening of data # convert integer to day-of-year values
phenology <- phenology |>
  mutate(
    value = ifelse(value > 32656, NA, value),
    value = as.numeric(format(as.Date("1970-01-01") + value, "%j")),
    value = ifelse (value < 200, value, NA)
  )
```

```{r}
# convert tidy data into geospatial (terra SpatRast format)
phenology_raster <- MODISTools::mt_to_terra(
  phenology,
  reproject = TRUE
)
```

```{r}
# crop the DEM correspond to the size of the coverage of the MODIS phenology
dem <- terra::crop(
  x = dem,
  y = phenology_raster
)
```

```{r}
# grid of DEM and MODIS data do not align -> resampling of data to a common grid required
# grid of the highest resolution data as a template for this resampling, 
# taking the average across the extent of a MODIS pixel
# resample the dem using
# the mean DEM value in a
# MODIS pixel
dem <- terra::resample(
  x = dem,
  y = phenology_raster,
  method = "average"
)
```

```{r}
# mask the locations which
# have no data
dem <- terra::mask(
  dem,
  is.na(phenology_raster),
  maskvalues = TRUE
)
```

for interpretation of results: useful to look at different responses by land 
cover class
```{r}
# therefore only download MODIS land cover product 2012
# download and save land cover data
land_cover <- MODISTools::mt_subset(
  product = "MCD12Q1",
  lat = 46.6756,
  lon = 7.85480,
  band = "LC_Type1",
  start = "2012-01-01",
  end = "2012-12-31",
  km_lr = 100,
  km_ab = 100,
  site_name = "swiss",
  internal = TRUE,
  progress = FALSE
)
```

```{r}
# now, convert to geospatial format as before
land_cover_raster <- MODISTools::mt_to_terra(
  land_cover,
  reproject = TRUE
)
```

phenology trends:

```{r}
## trends in phenology in relation to topography (plotting data side by side)
p <- ggplot() +
  tidyterra::geom_spatraster(data = dem) +
  scale_fill_viridis_c(
    na.value = NA,
    name = "altitude (m)"
  ) +
  theme_bw()

p2 <- ggplot() +
  tidyterra::geom_spatraster(data = phenology_raster) +
  scale_fill_viridis_c(
    na.value = NA,
    name = "DOY"
  ) +
  theme_bw()
```

```{r}
# compositing with patchwork package # plot showing both the DEM and MODIS
library(patchwork)
p + p2 + 
  plot_layout(ncol = 1) + 
  plot_annotation(
    tag_levels = "a",
    tag_prefix = "(",
    tag_suffix = ")"
  )
```

plot relation between topography and start of season (phenology) across the scene
```{r}
# non-spatially
# scatterplot of MODIS in function of altitude
# white line -> regression fit to all available data
# convert to data frame and merge
dem_df <- as.vector(dem)
phenology_df <- as.vector(phenology_raster)
sct_df <- data.frame(
  altitude = dem_df,
  doy = phenology_df
)

ggplot(
  data = sct_df,
  aes(
    altitude,
    doy
  )
) +
  geom_hex() +
  scale_fill_viridis_c(trans="log10") +
  geom_smooth(
    method = "lm",
    se = FALSE,
    colour = "white",
    lty = 2
  ) +
  labs(
    x = "altitude (m)",
    y = "MODIS vegetation greenup (DOY)"
  ) +
  theme_bw()
```

```{r}
# the fit linear regression above has the following parameters
# fit a linear regression to the data of the figure above
# (for the pre-processing see the collapsed code of the figure)
fit <- lm(doy ~ altitude, data = sct_df)
print(summary(fit))
```

```{r}

```

```{r}

```

```{r}

```

```{r}

```