---
title: "Report Exercise: Phenology Modelling"
author: "Salome Hayler"
date: "22.12.2025"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    self_contained: true
---

```{r setup, include=FALSE}
options(
  repos = c(CRAN="https://cloud.r-project.org")
)
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
rm(list = ls()) # clear environment
```

# 1. Introduction

This report is about with phenology modeling in the Boston region (in the northeastern United States) using data from MODIS, the PhenoCam network, and DAYMET climate data from Harvard Forest (see Chapter 6). Phenology modeling is important for understanding the effects of climate change (temperature increase and effects on, for example, earlier leaf emergence in plants) and therefore for providing models that are as accurate as possible. One approach to achieving this is to use growing degree days (GDD). GDD is defined as the cumulative sum of temperatures above a certain threshold, usually T₀ = 5 °C. This report discusses three possible ways to improve the original phenology model (from the book “A Handful of Pixels” by Koen Hufkens (2025) https://geco-bern.github.io/handfull_of_pixels/). The simplest approach is implemented as part of this exercise to accumulate GDD only after at least three consecutive days above the growth degree threshold of 5 °C, as plants need a stable period of warmth over several days to initiate leaf growth (see references and further explanations in Chapter 2.3). The model results are then compared with the MODIS MCD12Q2 phenology product.

# 2. Ideas of model improvement

**Accumulate GDD only after consecutive warm days (>5 °C)**

Classical GDD approaches accumulate heat on each day that the threshold is exceeded, ignoring previous conditions. Many plants need several consecutive warm days to end their dormancy. In this model, the threshold is set to at least three consecutive days with temperatures above 5 °C. This reduces the risk of counting individual warm days followed by frost. No specific value for the number of consecutive warm days could be found in the literature, and this probably varies between species. For this reason, I only use a number of at least 3 days to define a rule for consecutive warm days.

**Dynamic thresholds and adaptive GDD approaches for different locations**

Fixed base temperatures (e.g., 5 °C) and fixed GDD thresholds may not work equally well in different climate zones, for different species, or in different hemispheres. In a global model, thresholds and heat accumulation rules should take into account differences in growing seasons between the northern and southern hemispheres, as well as local responses of species to seasonal effects. Adjusting thresholds or accumulation rules based on local seasonal patterns could improve the model's universality across regions. For a truly global model, many other factors (that influence plant phenology) would also have to be taken into account. In addition to temperature, I am thinking of influences such as water availability, soil conditions, altitude, distance from the coast, wind conditions, and local microclimate (e.g., solar radiation or shade).

**Use longer data series and consistent data sets for calibration**

Calibrating the model using a single year (e.g., 2010) can lead to biased results, as extreme years are not representative of typical conditions. Using longer time series would allow for more general statements about the timing of leaf emergence, capture variability between years, and account for changes in temperature patterns over multiple years, including trends caused by climate change. Consistent use of the same datasets for calibration and evaluation, such as DAYMET and MODIS, would also improve spatial predictions and reduce discrepancies between years.

# 3. Initial model
This chapter reproduces the phenology modeling workflow presented in Chapter 6 of “A Handful of Pixels” without modification. The code has been taken directly from the book to understand how the phenology model is constructed and to create a basic growing degree day (GDD) model that serves as a reference for subsequent model improvement. If you don't want to look at the first part of the model, skip ahead to Chapter 4 (Model improvement).

I will use the phenocamr package which interfaces with the phenocam network API
to download time series of vegetation greenness and derived phenology metrics.
```{r }
# required packages
packages <- c(
  "dplyr",
  "ggplot2",
  "readr",
  "patchwork",
  "phenocamr",
  "GenSA",
  "terra",
  "leaflet",
  "tidyterra",
  "leaflet",
  "MODISTools",
  "appeears", # required if download DAYMET data from NASA
  "keyring",  # required if download DAYMET data from NASA
  "sf",       # required if download DAYMET data from NASA
  "here",
  "knitr",
  "rmarkdown"
)
```

```{r, include=FALSE}
# install missing packages
for (pkg in packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
}

# load frequently used packages
library(dplyr)     # for data manipultations
library(ggplot2)   # for making plots
library(phenocamr) # for download + processing of PhenocamNetwork data
library(terra)     # for processing raster data
library(patchwork) # for combining multiple plots
library(leaflet)   # for creating interactive web maps

# set project root
here::i_am("agds_report_salomehayler.Rproj")
```

Download greenness time series, calculate phenology (phenophases), amend with DAYMET data:
```{r download data, message=FALSE}
phenocamr::download_phenocam(
  site = "harvard$",
  veg_type = "DB",
  roi_id = "1000",
  daymet = TRUE,
  phenophase = TRUE,
  trim = 2022,
  out_dir = tempdir()
  )
```

```{r, message=FALSE}
harvard_phenocam_data <- readr::read_csv(
  file.path(tempdir(), "harvard_DB_1000_3day.csv"), 
  comment = "#"
  )
```

```{r, message=FALSE}
# reading in harvard phenology only retaining
# spring (rising) phenology for the GCC 90th
# percentile time series (the default)
harvard_phenology <- readr::read_csv(
  file.path(
    tempdir(),
    "harvard_DB_1000_3day_transition_dates.csv"
    ),
  comment = "#"
) |>
  dplyr::filter(
    direction == "rising",
    gcc_value == "gcc_90"
  )
```

Plot time series of the 90th percentile 3-daily green chromatic coordinate GCC:
```{r plot, fig.cap="Figure 1: Time series of the 90th percentile 3-daily green chromatic coordinate (GCC) values for the Harvard forest EMS tower. PhenoCam GCC data are shown by the dark grey line, with threshold based (25% seasonal amplitude) derived phenology marked as black dots."}
ggplot(harvard_phenocam_data) +
  geom_line(
    aes(
      as.Date(date),
      smooth_gcc_90
    ),
    colour = "grey25"
  ) +
  geom_point(
    data = harvard_phenology,
    aes(
      as.Date(transition_25),
      threshold_25
    )
  ) +
  labs(
    x = "",
    y = "GCC"
  ) +
  theme_bw() +
  theme(
    legend.position = "none"
  )
```

Preprocess data:
```{r }
# growing degree days = cumulative sum of temperatures above a specified threshold, mostly T0=5°C
# return mean daily temperature as well
# as formal dates (for plotting)
harvard_temp <- harvard_phenocam_data |>
  group_by(year) |>
  dplyr::mutate(
    tmean = (tmax..deg.c. + tmin..deg.c.)/2
  ) |> 
  dplyr::mutate(
    date = as.Date(date),
    gdd = cumsum(ifelse(tmean >= 5, tmean - 5, 0))
  ) |>
  dplyr::select(
    date,
    year,
    tmean,
    gdd
  ) |>
  ungroup()
```

```{r }
# convert the harvard phenology data and only retain required data
harvard_phenology <- harvard_phenology |>
  mutate(
    doy = as.numeric(format(as.Date(transition_25),"%j")),
    year = as.numeric(format(as.Date(transition_25),"%Y"))
  ) |>
  select(
    year,
    doy,
    transition_25,
    threshold_25
  )
```

Now: Plot the temperature series of all days (distinguish between days where temperature threshold is exceeded or not).
```{r }
# grab only the 2010 value of spring phenology
harvard_phenology_2010 <- harvard_phenology |>
  dplyr::filter(
    year == 2010
  )

harvard_gdd_value <- harvard_temp |>
  dplyr::filter(
    date == harvard_phenology_2010$transition_25
  )

p <- ggplot(harvard_temp) +
  geom_line(
    aes(
      date,
      tmean
    )
  ) +
  geom_point(
    aes(
      date,
      tmean,
      colour = tmean > 5,
      group = 1
    )
  ) +
  geom_vline(
    data = harvard_phenology_2010,
    aes(
      xintercept = as.Date(transition_25)
    )
  ) +
  scale_colour_discrete(
    type = c(
      "blue",
      "red"
    )
  ) +
  labs(
    x = "",
    y = "Temperature (deg. C)"
  ) +
  xlim(
    c(
      as.Date("2010-01-01"),
      as.Date("2010-06-30")
    )
  ) +
  theme_bw() +
  theme(
    legend.position = "none"
  )
```

```{r }
p2 <- ggplot(harvard_temp) +
  geom_line(
    aes(
      date,
      gdd
    )
  ) +
  geom_point(
    aes(
      date,
      gdd,
      colour = tmean > 5,
      group = 1
    )
  ) +
  scale_colour_discrete(
    type = c(
      "blue",
      "red"
    )
  ) +
  geom_vline(
    data = harvard_phenology_2010,
    aes(
      xintercept = as.Date(transition_25)
    )
  ) +
  geom_hline(
    data = harvard_gdd_value,
    aes(
      yintercept = gdd
    ),
    lty = 2
  ) +
  labs(
    x = "",
    y = "GDD (deg. C)"
  ) +
  xlim(
    c(
      as.Date("2010-01-01"),
      as.Date("2010-06-30")
    )
  ) +
  ylim(c(0, 1000)) +
  theme_bw()  +
  theme(
    legend.position = "none"
  )
```

```{r, warning=FALSE, fig.cap="Figure 2: a) Temperature time series from Jan-Jul showing daily mean temperature below 5°C in blue and those above in red b) Growing degree days (GDD) as the cummulative temperature sum of warm (red) days; the black lines mark the observed leaf-out-day in spring, and the corresponding GDD, respectively." }
# compositing
p + p2 + 
  plot_layout(ncol = 1) + 
  plot_annotation(
    tag_levels = "a",
    tag_prefix = "(",
    tag_suffix = ")"
  )
```

The GDD increases on days when temperatures exceed the threshold during the growing season, and in 2010, leaf emergence in spring was observed on day 114 of the year (DOY), corresponding to a GDD of 130.44 °C. While this value provides a simple estimate for leaf emergence in a single year, it does not necessarily yield accurate predictions across multiple years or locations, necessitating model generalization.

Growing degree day model optimization to generalize the model: 
```{r }
# function: temp series as first argument, two parameters  (temp threshold above which temperatures are accumulated & critical GDD that determines the DOY at which leaf-out is predicted)
gdd_model <- function(temp, par) {
  # split out parameters from a simple
  # vector of parameter values
  temp_threshold <- par[1]
  gdd_crit <- par[2]
  
  # accumulate growing degree days for
  # temperature data
  gdd <- cumsum(ifelse(temp > temp_threshold, temp - temp_threshold, 0))
  
  # figure out when the number of growing
  # degree days exceeds the minimum value
  # required for leaf development, only
  # return the first value
  doy <- unlist(which(gdd >= gdd_crit)[1])
  
  return(doy)
}
```

```{r }
# confirm that the model function
# returns expected results (i.e. DOY 114)
# (we filter out the year 2010, but
# removing the filter would run the
# model for all years!)
prediction <- harvard_temp |>
  dplyr::filter(
    year == 2010
  ) |>
  group_by(year) |>
  summarize(
    pred = gdd_model(
      temp = tmean,
      par = c(5, 130.44)
    )  
  )
```

```{r }
print(prediction) # a tibble showing the predicted GDD at which the leaf-out was observed
```

Here, the critical GDD value was taken directly from the observed leaf-out-date, resulting in a model that accurately reproduces the observed leaf-out time.

Although the predicted leaf-out-date matches the observation, it remains unclear whether this parameter selection is valid beyond a single year or a single location. To clarify this question, the phenology model is calibrated by making predictions over a longer period of time and comparing them with observations.
```{r }
# goal: minimize the error (cost function) between model results (for 
# given set of parameters (temp & critical GDD) and our observed data 
# estimating parameters efficiently 
## nls()function (square error minimalization for any (non-linear) function)
## or simulated annealing GenSA package
## or Bayesian Tools package
# run model and compare to true values
# returns the RMSE (between observed and predicted values)
rmse_gdd <- function(par, data) {
  
  # split out data
  drivers <- data$drivers
  validation <- data$validation
  
  # calculate phenology predictions
  # and put in a data frame
  predictions <- drivers |>
    group_by(year) |>
    summarise(
      predictions = gdd_model(
        temp = tmean,
        par = par
      )
    )
  
  predictions <- left_join(predictions, validation, by = "year")
  
  rmse <- predictions |>
    summarise(
      rmse = sqrt(mean((predictions - doy)^2, na.rm = TRUE))
    ) |>
    pull(rmse)
  
  # return rmse value
  return(rmse)
}
```

```{r }
# often, starting model parameters and limits to the parameter space are required (temp. threshold and physiological limits of plant activity)
# starting model parameters
par = c(0, 130)
```

```{r }
# limits to the parameter space
lower <- c(-10,0)
upper <- c(45,500)
```

```{r }
# data needs to be provided in a consistent
# single data file, a nested data structure
# will therefore accept non standard data formats
data <- list(
  drivers = harvard_temp,
  validation = harvard_phenology
)
```

Optimize the model parameters:
```{r }
optim_par = GenSA::GenSA(
  par = par,
  fn = rmse_gdd,
  lower = lower,
  upper = upper,
  control = list(
    max.call = 4000 # calling cost function 4000 times
  ),
  data = data
)$par
```

Run the model for all years (plug the optimized parameters back into the model)
```{r }
# to get the phenology predictions
predictions <- harvard_temp |>
  group_by(year) |>
  summarize(
    prediction = gdd_model(
      temp = tmean,
      par = optim_par
    )  
  )
```

```{r, message=FALSE, warning=FALSE, fig.cap="Figure 3: Scatterplot of observed versus predicted leaf-out dates (DOY), with a linear regression shown as a solid dark grey line and its confidence intervals indicated by a light grey shaded area." }
# show results visually
# join predicted with observed data
validation <- left_join(predictions, harvard_phenology)

ggplot(validation) +
  geom_smooth(
    aes(
      doy,
      prediction
    ),
    colour = "grey25",
    method = "lm"
  ) +
  geom_point(
    aes(
      doy,
      prediction
    )
  ) +
  geom_abline(
    intercept=0, 
    slope=1, 
    linetype="dotted"
  ) +
  labs(
    x = "Observed leaf-out date (DOY)",
    y = "Predicted leaf-out date (DOY)"
  ) +
  theme_bw()  +
  theme(
    legend.position = "none"
  )

```

# 4.  Model improvement
This chapter improves the original GDD model by implementing a rule that only accumulates growing degree days during periods of at least three consecutive days with temperatures above 5 °C.

```{r}
# improved GDD model function -> only starts counting GDD when there are at least 3 days in a row above mean temperature of 5°C
gdd_model_optim <- function(temp, par, min_run = 3) {
  # temp: numeric vector of daily mean temperatures
  # par:  c(temp_threshold, gdd_crit)
  # min_run: number of consecutive warm days required before counting GDD
  
  temp_threshold <- par[1]   # temperature threshold (°C)
  gdd_crit <- par[2]         # total GDD required for leaf-out
  
  # identify days warmer than the threshold
  warm_days <- temp > temp_threshold
  
  # find runs of consecutive warm days using rle()
  run_lengths <- rle(warm_days)
  
  # only periods where at least min_run (= at least 3) consecutive warm days are retained as TRUE
  valid_runs <- rep(run_lengths$values & run_lengths$lengths >= min_run,
                    run_lengths$lengths)
  
  # accumulate GDD only during ≥3-day warm periods
  gdd <- cumsum(ifelse(valid_runs, temp - temp_threshold, 0))
  
  # first day on which GDD reaches the critical value
  doy <- which(gdd >= gdd_crit)[1]
  
  return(doy)
}
```

```{r}
# RMSE cost function for the optimized model (gddmodel_optim)
# for measuring how well the model predictions match observed phenology dates

rmse_gdd_optim <- function(par, data, min_run = 3) {
  # par:  c(temp_threshold, gdd_crit)
  # data: list with elements:
  
  drivers <- data$drivers       # daily temperature data
  validation <- data$validation # observed phenology data
  
  # predict phenology for each year
  predictions <- drivers %>%
    dplyr::group_by(year) %>%
    dplyr::summarise(
      prediction = gdd_model_optim(tmean, par, min_run = min_run),
      .groups = "drop"
    )
  
  # join predicted with observed data
  merged <- dplyr::left_join(validation, predictions, by = "year")
  
  # calculate root mean squared error (RMSE)
  rmse <- sqrt(mean((merged$prediction - merged$doy)^2, na.rm = TRUE))
  
  return(rmse)
}
```

```{r}
# set starting parameters and limits to the parameter space (similar as for the initial model)
par <- c(0, 130)
lower <- c(-10, 0)
upper <- c(45, 500)

# data required as list structure
data <- list(
  drivers = harvard_temp,
  validation = harvard_phenology
)
```

Now to the parameter optimization:
```{r}
# parameter optimization with GenSA
optim_par <- GenSA::GenSA(
  par = par,
  fn = function(par, data) rmse_gdd_optim(par, data, min_run = 3),
  lower = lower,
  upper = upper,
  control = list(max.call = 4000), # calls the cost function 4000 times (as also done for the initial model)
  data = data
)$par
```

Now run the model for all years and put the optimized parameters back into the model:
```{r}
# --- Predict phenology using optimized parameters ---
predictions <- harvard_temp %>%
  group_by(year) %>%
  summarise(
    prediction = gdd_model_optim(tmean, par = optim_par, min_run = 3),
    .groups = "drop"
  )
```

Show results visually: 
```{r, message=FALSE}
# join predicted with observed data
validation_optimmodel <- left_join(predictions, harvard_phenology, by = "year")
```

```{r, message=FALSE, warning=FALSE, fig.cap="Figure 4: Scatterplot (for the optimised model) of observed versus predicted leaf-out dates (DOY), with a linear regression shown as a solid dark grey line and its confidence intervals indicated by a light grey shaded area."}
# plot predicted vs observed leaf-out dates
ggplot(validation_optimmodel) +
  geom_smooth(
    aes(x = doy, y = prediction),
    colour = "grey25",
    method = "lm"
  ) +
  geom_point(aes(x = doy, y = prediction)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted") +
  labs(
    x = "Observed leaf-out date (DOY)",
    y = "Predicted leaf-out date (DOY)",
    title = "Optimized GDD Model (at least 3-days above 5°C)"
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

This model only uses average daily temperatures, so it does not record absolute hourly temperatures. As a result, night frosts can still occur even when the average daily temperature is above 5 °C, which could have an impact on plant phenology in reality. In addition, three consecutive warm days may not be sufficient for all plants, as frost after these days can still damage or kill buds.

## 4.1 Comparison of the RMSE (initial vs. optimised model)
To quantify how well the models reproduce the observed leaf-out dates, the root mean squared error (RMSE) is calculated for both the original GDD model and the optimized GDD model, providing a measure of the prediction error between the observed and modeled DOY values.

```{r}
# RMSE comparison using the rmse functions from both models
rmse_initial <- rmse_gdd(par = optim_par, data = data)
rmse_optimized <- rmse_gdd_optim(par = optim_par, data = data, min_run = 3)

# show results in a tibble
rmse_comparison <- tibble(
  Model = c("Initial GDD Model", "Optimized GDD Model"),
  RMSE = c(rmse_initial, rmse_optimized)
)

print(rmse_comparison)
```

The original GDD model has an RMSE of 4.70 days, while the optimized GDD model achieves a slightly lower RMSE of 4.32 days. This indicates a slight improvement in the fit, although the difference is very small and barely noticeable in the scatter plots. Both the original and optimized GDD models show a clear positive relationship between the observed and predicted leaf-out-dates. The points are roughly aligned along the 1:1 line, suggesting that the models generally capture the timing of leaf-out. Comparing the two scatterplots (Figure 3 and Figure 4), the optimized model with the 3-day threshold appears to predict slightly earlier leaf-out-date for some of the observed data in the middle range (approximately DOY 125–130), as the fitted line lies just above the points for earlier observations. For the later leaf-out-dates (DOY > 130), the predictions are very similar to those of the original model. The slight improvement is to be expected, as only a simple change (3 consecutive warm days) was made.

## 4.2 Spatial scaling of the improved model
For the spatial scaling of the optimized GDD model, the calibrated parameters were applied to a larger area around Boston in the eastern United States (longitude: -72 to -70, latitude: 42 to 44). The model is based on DAYMET grid data, specifically the daily minimum and maximum temperatures, which are averaged to calculate the daily average temperatures in accordance with the calibration data. The temperature data was previously downloaded using the AppEEARS R package and is stored locally in the “data_raw” folder. Since the data is already available, no additional NASA account setup or download is required to perform the spatial analysis. The R Markdown script (not shown in the rendered document) contains additional code blocks that illustrate how the original data was accessed (see https://bluegreen-labs.github.io/appeears/articles/appeears_vignette.html for more information).

```{r, eval=FALSE, include=FALSE }
## in the following code chunks: Instructions for downloading DAYMET raster data using your own account -> this allows the model to be applied across a larger landscape
# store NASA Earth Data passwort (runs only once)
key_set(service = "your username NASA")  
# -> enter your NASSA Earth passwort when promted

# retrieve stored passwort for use in the skript
my_password <- key_get(service = "your username NASA")
my_password  # verify that password was stored correctly
```

```{r, eval=FALSE, include=FALSE }
rs_set_key(user = "your username NASA",     
           password = "your passwort NASA"  
)
```

```{r, eval=FALSE, include=FALSE }
# to see whether the setup has worked, you can try to get the 
# metadata of the product DAYMET v4:
rs_layers("DAYMET.004")
```

```{r, eval=FALSE, include=FALSE }
# define the bounding box coordinates for the Boston area
xmin <- -72; xmax <- -70
ymin <- 42; ymax <- 44
coords <- matrix(c(
  xmin, ymin,
  xmax, ymin,
  xmax, ymax,
  xmin, ymax,
  xmin, ymin  # closing the polygon by repeating first point
), ncol = 2, byrow = TRUE)
```

```{r, eval=FALSE, include=FALSE }
# create a simple feature geometry collection with CRS
roi <- st_sf(geometry = st_sfc(st_polygon(list(coords)), 
                               crs = 4326)  # WGS 84)
)
```

```{r, eval=FALSE, include=FALSE }
# build the area based request/task
df <- data.frame(
  task = "DAYMET.004_2012",
  subtask = "subtask",
  latitude = NA,
  longitude = NA,
  start = "2012-01-01",
  end = "2012-12-31",
  product = "DAYMET.004",
  layer = c("tmax", "tmin")
)
```

```{r, eval=FALSE, include=FALSE }
task <- rs_build_task(
  df = df,
  roi = roi,
  format = "netcdf4"
)
```

```{r, eval=FALSE, include=FALSE }
# request the task to be executed (takes some time -> see https://appeears.earthdatacloud.nasa.gov/explore)
rs_request(
  request = task,
  user = "your username NASA",
  transfer = TRUE,
  path = here::here("data_raw"),
  verbose = TRUE
)
```

Load the previously downloaded NASA data from the folder data_raw
```{r }
# Load the downloaded daily data:
r1 <- terra::rast(
  here::here("data_raw",
             "DAYMET.004_2012", 
             "DAYMET.004_1km_aid0001.nc"), 
  guessCRS=FALSE)
```

```{r }
# Assign the correct CRS, e.g. WGS84 (EPSG:4326)
terra::crs(r1) <- "epsg:4326"
```

The variables can be accessed with:
r1["tmax"]
r1["tmin"]

Process downloaded data:
```{r }
# calculate the daily mean values based on 'tmin' and 'tmax'
mean_layer <- terra::mean(r1["tmax"], 
                          r1["tmin"])
```

```{r }
# fix the variable naming
terra::varnames(mean_layer) <- "tmean"                             
names(mean_layer) <- gsub("tmax","tmean",names(mean_layer))
```

Showing the result (a raster with tmean for each day) visually:
```{r, fig.cap="Figure 5: Mean daily temperatures (first 16 days of 2012) in the greater Boston area."}
# plot: result is raster containing a single variable "tmean" for each day
terra::plot(mean_layer)
```

```{r }
# reduce memory footprint of the calculations
# subset to first 180 days
ma_nh_temp <- terra::subset(
  mean_layer,
  1:180
)
```

Now, apply the optimized model to this temperature raster
```{r }
# apply model to this raster (cube)
predicted_phenology_optim <- terra::app( 
  ma_nh_temp,
  fun = gdd_model_optim, 
  par = optim_par
)
```

The result is an interactive map of the spatially scaled optimized growing degree model using DAYMET daily mean temperature data
```{r }
# set te colour scale manually
pal <- colorNumeric(
  "magma",
  terra::values(predicted_phenology_optim), # ...optim geändert
  na.color = "transparent"
)
```

```{r, fig.cap="Figure 6: Interactive map of the spatially scaled optimized growing degree model using DAYMET daily mean temperature data, covering the greater Boston area in the southeast up to the White Mountains in the northwest." }
# build the leaflet map
# using ESRI tile servers
# and the loaded demo raster
leaflet() |> 
  addProviderTiles(providers$Esri.WorldImagery, group = "World Imagery") |>
  addProviderTiles(providers$Esri.WorldTopoMap, group = "World Topo") |>
  addRasterImage(
    predicted_phenology_optim,
    colors = pal,
    opacity = 0.8,
    group = "Phenology model results"
  ) |>
  addLayersControl(
    baseGroups = c("World Imagery","World Topo"),
    position = "topleft",
    options = layersControlOptions(collapsed = FALSE),
    overlayGroups = c("Phenology model results")
  ) |>
  addLegend(
    pal = pal,
    values = terra::values(predicted_phenology_optim),
    title = "DOY")
```

```{r}
# DOY statistis for the optimized model over the Boston region (mean, min, max)
mean_DOY_optim <- terra::global(predicted_phenology_optim, fun = mean, na.rm = TRUE)
min_DOY_optim <- terra::global(predicted_phenology_optim, fun = min, na.rm = TRUE)
max_DOY_optim <- terra::global(predicted_phenology_optim, fun = max, na.rm = TRUE)

mean_DOY_optim
min_DOY_optim
max_DOY_optim
```

# 5. Comparison of the model output with the MODIS MCD12Q2 phenology product

To evaluate the performance of the optimized GDD model on a larger spatial scale, its predictions are compared with the MODIS MCD12Q2 land surface phenology product, which provides satellite-based data on vegetation start dates for the greater Boston area (latitude 42–44° N, longitude -72 to -70° E -> I used the mean value in each case).

## 5.1 MODIS MCD12Q2 for Boston area
``` {r}
# download and save phenology data
phenology_boston <- MODISTools::mt_subset(
  product = "MCD12Q2",
  lat = 43,  # coordinate for Boston area
  lon = -71, # coordinate for Boston area
  band = "Greenup.Num_Modes_01",
  start = "2012-01-01",
  end = "2012-12-31",
  km_lr = 100,
  km_ab = 100,
  site_name = "swiss",
  internal = TRUE,
  progress = FALSE
)
```

First, MODIS requires preprocessing: The product documentation also shows that phenological metrics are counted as days starting from January 1, 1970. To simplify interpretation, we will convert these integer values, counted from 1970, into values for the day of the year (using as.Date() and format()). We only consider phenological events in the first 200 days of the year, as we are focusing on spring. Later dates are most likely incorret.

```{r}
# screening of data 
# convert integer to day-of-year values
phenology_boston <- phenology_boston |>
  mutate(
    value = ifelse(value > 32656, NA, value),
    value = as.numeric(format(as.Date("1970-01-01") + value, "%j")),
    value = ifelse (value < 200, value, NA)
  )
```

```{r}
# convert tidy data into geospatial (terra SpatRast format)
phenology_raster_boston <- MODISTools::mt_to_terra(
  phenology_boston,
  reproject = TRUE
)
```

Now, plot the MODIS data:
```{r, fig.cap = "Figure 7: MODIS land surface phenology for the greater Boston area as the Day-of-year (DOY) in 2012."}
ggplot() +
  tidyterra::geom_spatraster(data = phenology_raster_boston) +
  scale_fill_viridis_c(
    na.value = NA,
    name = "DOY"
    ) +
  theme_bw()
``` 

```{r }
# DOY statistics for the MODIS raster over Boston (mean, min, max)
mean_DOY_MODIS <- terra::global(phenology_raster_boston, fun = mean, na.rm = TRUE)
min_DOY_MODIS <- terra::global(phenology_raster_boston, fun = min, na.rm = TRUE)
max_DOY_MODIS <- terra::global(phenology_raster_boston, fun = max, na.rm = TRUE)

mean_DOY_MODIS
min_DOY_MODIS
max_DOY_MODIS
```

The MODIS MCD12Q2 product provides phenological transition data derived from time series of vegetation indices. The greening data is determined based on the increase in greenness of the canopy rather than on direct physiological processes. Therefore, the product reflects responses at the tree canopy level that are integrated over relatively large spatial areas (500 m resolution) (for more information, see: https://cran.r-project.org/web/packages/MODISTools/index.html).

## 5.2 Comparing the data spatially
The spatial patterns of spring greening derived from the optimized GDD model and the MODIS MCD12Q2 product are highly consistent on a broad scale. Both datasets show earlier greening in the south and along the coasts of the Boston region and increasingly later greening toward the interior and higher elevations. This agreement suggests that both approaches capture the dominant influence of temperature on spring phenology.

Despite this general agreement, the two datasets differ in their statistical range. For the Boston region, the MODIS product shows a mean greening date of DOY 101 with values ranging from DOY 67 to DOY 141. In contrast, the optimized GDD model predicts a later mean greening date of DOY 111 with a narrower range from DOY 101 to DOY 147. This suggests that, compared to MODIS, the model generally predicts later and spatially more homogeneous greening dates. Neither dataset reproduces the DOY of approximately 114 obtained from the original point-based model, which reflects site-level conditions in 2010 rather than regional-level phenology in 2012.

The higher spatial variability of the MODIS product reflects its sensitivity to land cover heterogeneity within 500-m pixels, including mixtures of forest, urban areas, and agricultural land. The optimized GDD model produces smoother spatial patterns because it is based purely on gridded temperature data and assumes a uniform temperature response across the region.

Since MODIS provides a phenological metric derived from satellite data and not direct observations of physiological leaf emergence, an RMSE comparison between MODIS and the model is not meaningful. Model performance was therefore previously evaluated solely on the basis of PhenoCam observations.

## 5.3 Reasons for similarities and differences in patterns
Several factors explain the remaining differences between the optimized GDD model and the MODIS MCD12Q2 product. First, the two approaches describe different aspects of phenology. The GDD model estimates the timing of physiological leaf-out based on accumulated temperature, while MODIS captures canopy greening based on changes in surface reflectance. As a result, MODIS can detect greening earlier than the temperature-based thresholds are reached.

Second, the datasets differ in their spatial resolution. MODIS pixels integrate vegetation signals over large areas, resulting in higher spatial variability, while the GDD model reflects the smoother spatial structure of DAYMET temperature data.

Finally, the comparison is affected by temporal inconsistencies. The optimized GDD model parameters were calibrated using PhenoCam observations from 2010, while both the spatially scaled model outputs and the MODIS product reflect conditions in 2012. The optimized model therefore applies parameters influenced by 2010 conditions to a different climate year, contributing to systematic differences in the predicted timing of greening.

## 5.4 2010 as a special year for the northeastern US
The year 2010 was a climatologically exceptional period for the northern hemisphere, marked by record cold temperatures in the winter of 2009/2010. These extreme conditions were associated with strongly negative phases of the North Atlantic Oscillation, which led to prolonged cold air outbreaks over North America (Wang et al., 2010).

The original GDD model and the optimized model parameters are based on PhenoCam observations from 2010, which means that the estimated temperature thresholds implicitly reflect the phenological responses after this unusually cold winter. Such conditions can delay the accumulation of growing degree days and postpone the leaf-out-date to later in the year, contributing to the later DOY values of the original model.

However, the spatial predictions of the optimized GDD model were created using DAYMET temperature data from 2012, and the MODIS MCD12Q2 product used for comparison also represents the phenology in 2012. Consequently, neither the optimized model nor the MODIS dataset reflect the exceptional winter conditions of 2009/2010.

This temporal discrepancy explains why the spatially leaf-out respectively greening data from both datasets (my optimised model, respectively the MODIS data) deviate from the DOYs derived from the original model based on 2010. The comparison presented here therefore reflects the typical phenological patterns for 2012 rather than the extreme conditions observed in the winter of 2009/2010.

# 6. References & Declaration of Authorship
I hereby declare that I (Salome Hayler) have written this report myself. It is based on fundamental codes and general information from the book “A Handful of Pixels” by Koen Hufkens (2025), general information from the book “Applied Geodata Science (v1.0)” by Benjamin Stocker et. al (2023), and information from lectures (by Prof. Dr. Benjamin Stocker & Dr. Fabian Bernhard). All other sources of information are listed in this section and have been cited in the text. 

AI (ChatGPT) was used to solve technical problems that could not be solved otherwise. For this report, this particularly concerned the download of DAYMET data, as the instructions from the appears package did not work at all in the beginning. 

I used DeepL (free version) to check the grammatical correctness of some passages and improve them in terms of English scientific language.

Last but not least, “R Markdown: The Definitive Guide” by Yihui Xie, J. J. Allaire, and Garrett Grolemund (2023) (available at: https://bookdown.org/yihui/rmarkdown/) served as a helpful guide to Markdown syntax.

**Data** 
(see also README in folder data_raw)

*PhenoCam data:* Observations from the Harvard PhenoCam website were used to obtain time series of tree crown greenness (GCC) and derive leaf-out data for model calibration and validation.

*DAYMET data:* Daily gridded temperature data (Tmax and Tmin) were obtained from NASA DAYMET to calculate daily average temperatures and accumulate growing degree days (GDD) for each year and across the entire spatial domain.

*MODIS MCD12Q2:* The MODIS Land Surface Phenology product was used for spatial comparison of greening data. These phenological measurements derived from satellite data provide an independent reference for evaluating the spatial patterns predicted by the GDD model.

**Additional Literature for information about the year 2010**
Wang, C. et al. (2010). The record-breaking cold temperatures during the winter of 2009/2010 in the Northern Hemisphere. Atmospheric Science Letters. https://doi.org/10.1002/asl.278.