---
title: "Report Exercise: Digital Soil Mapping"
author: "Salome Hayler"
date: "22.12.2025"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
options(
  repos = c(CRAN="https://cloud.r-project.org")
)
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
rm(list = ls()) # clear environment
```

# 1. Introduction
A lot of different factors influence soil conditions, mostly influenced by XXX (formel soil)
soil formula hinschreiben und soil factors nennen
It is of interest to map soil conditions over larger areas
climate change, water conditions
This tutorial shows a digital soil mapping flow for the categorial variable waterlog.100 (explanation variable) -> variable erklären was die ist und was anzeigt?
The data used includes different variables, from which sampling locations around Bern?
Was ich mache in diesem Markdown und worum es geht
wieso ist thema wichtig?

# 2. Re-implemention of the digital soil mapping workflow for waterlog.100
The necessary data for this exercise was downloaded previously to the re-implementation of the workflow for the categorical variable waterlog.100. The full code for downloading and preprocessing the original data can be found in the R Markdown file and isn't included in the rendered document.

```{r}
# required packages
packages <- c(
  "dplyr",
  "tidyr",
  "readr",
  "tibble",
  "terra",
  "sf",
  "tidyterra",
  "ggplot2",
  "lattice",
  "ranger",
  "Boruta",
  "caret",
  "recipes",
  "pROC",
  "knitr",
  "rmarkdown",
  "here"
)
```

```{r, include=FALSE}
# install missing packages
for (pkg in packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
}

# load frequently used packages
library(dplyr)      # for data manipultations
library(tidyr)      # for data transformations
library(terra)      # for processing raster data
library(here)       # for reproducibility
library(ggplot2)    # for making plots
library(caret)      # for cross-validation, hyperparameter tuning
library(ranger)     # for random forest
library(Boruta)     # for variable selection
library(sf)         # for spatial features
library(pROC)       # for ROC/AUC
library(knitr)      # for rendering R Markdown
library(patchwork)  # for combining multiple plots
library(lattice)    # for intern plots e.g. during cross validation

# set project root
here::i_am("agds_report_salomehayler.Rproj")
```

```{r, eval=FALSE, include=FALSE}
# load soil data from sampling locations
df_obs <- readr::read_csv(
  here::here("data_raw/soildata/berne_soil_sampling_locations.csv")
  )

# display data
head(df_obs) |> 
  knitr::kable()
```

```{r, eval=FALSE, include=FALSE}
# load the covariates that we want to produce our soil maps with. These files are in the geoTIFF format - geolocated TIFF files.
# get a list with the path to all raster files
list_raster <- list.files(
  here::here("data_raw/geodata/covariates/"),
  full.names = TRUE
  )

# display data (lapply to clean names)
lapply(
  list_raster, 
  function(x) sub(".*/(.*)", "\\1", x)
  ) |> 
  unlist() |> 
  head(5) |> 
  print()
```

```{r, eval=FALSE, include=FALSE}
# load a raster file as example: Picking the slope profile at 2 m resolution
raster_example <- terra::rast(
  here::here("data_raw/geodata/covariates/Se_slope2m.tif")
  )
raster_example
```

```{r , eval=FALSE, include=FALSE}
# combine data: The {terra} package comes with the very useful tool to stack multiple rasters on top of each other if they share the spatial grid (extent and resolution). To do so, we just have to feed in the vector of file names list_raster
# load all files as one batch
all_rasters <- terra::rast(list_raster)
all_rasters
```

```{r , eval=FALSE, include=FALSE}
# extract coordinates from sampling locations
sampling_xy <- df_obs |> 
  dplyr::select(x, y)

# from all rasters, extract values for sampling coordinates
df_covars <- terra::extract(
  all_rasters,  # the raster we want to extract from
  sampling_xy,  # a matrix of x and y values to extract for
  ID = FALSE    # to not add a default ID column to the output
  )

df_full <- cbind(df_obs, df_covars)
head(df_full) |> 
  knitr::kable() 
```

```{r , eval=FALSE, include=FALSE}
vars_categorical <- df_covars |> 
  
  # get number of distinct values per variable
  dplyr::summarise(dplyr::across(dplyr::everything(), ~dplyr::n_distinct(.))) |> 
  
  # turn df into long format for easy filtering
  tidyr::pivot_longer(
    dplyr::everything(), 
    names_to = "variable", 
    values_to = "n"
    ) |> 
  
  # tilter out variables with 10 or less distinct values
  dplyr::filter(n <= 10) |>
  
  # extract the names of these variables
  dplyr::pull('variable')

cat("Variables with less than 10 distinct values:", 
    ifelse(length(vars_categorical) == 0, "none", vars_categorical))
```

```{r , eval=FALSE, include=FALSE}
# now that we have the names of the categorical values, we can mutate these columns in our data frame using the base function as.factor():
df_full <- df_full |> 
  dplyr::mutate(dplyr::across(all_of(vars_categorical), ~as.factor(.)))
```

```{r, eval=FALSE, include=FALSE}
# save the data
if (!dir.exists(here::here("data"))) system(paste0("mkdir ", here::here("data")))
saveRDS(df_full, 
        here::here("data/df_full.rds"))
```

## 2.1 Simple model for waterlog.100
explanation schreiben, was waterlog.100 für eine variable ist, also ob wassergesättigt oder nicht?? oder was sagt diese variable (im dokument nachschaun)

### 2.1.1 Random forest for waterlog.100
```{r, results='hide'}
# load dataframe
df_full_soil <- readRDS(here::here("data/df_full.rds"))

df_full_soil |> 
  knitr::kable()
```

```{r }
# specify target: binary waterlog.100
target_wl <- "waterlog.100"
```

```{r}
# make as factor
df_full_soil[[target_wl]] <- factor(df_full_soil[[target_wl]])
```

```{r }
# specify predictors_all: Remove soil sampling and observational data
predictors_all <- names(df_full_soil)[14:ncol(df_full_soil)]

cat("The target is:", target_wl,
    "\nThe predictors_all are:", paste0(predictors_all[1:8], sep = ", "), "...")
```

```{r }
# split dataset into training and testing sets
df_train_wl <- df_full_soil |> dplyr::filter(dataset == "calibration")
df_test_wl  <- df_full_soil |> dplyr::filter(dataset == "validation")
```

```{r }
# encode target variable as factor
df_train_wl[[target_wl]] <- factor(df_train_wl[[target_wl]]) 
df_test_wl[[target_wl]]  <- factor(df_test_wl[[target_wl]]) 
```

```{r }
# filter out any NA to avoid error when running a random forest
df_train_wl <- df_train_wl |> tidyr::drop_na()
df_test_wl  <- df_test_wl   |> tidyr::drop_na()
```

```{r }
# a little bit of verbose output
n_tot <- nrow(df_train_wl) + nrow(df_test_wl)

perc_cal <- (nrow(df_train_wl) / n_tot) |> round(2) * 100
perc_val <- (nrow(df_test_wl)  / n_tot) |> round(2) * 100

cat("For model training, we have a calibration / validation split of: ",
    perc_cal, "/", perc_val, "%")
```

```{r }
# ranger() crashes when using tibbles, so we are using the
# base R notation to enter the data
rf_basic_wl <- ranger::ranger( 
  y = df_train_wl[, target_wl],       # target variable
  x = df_train_wl[, predictors_all],  # Predictor variables
  seed = 42,                          # Specify seed for randomization to reproduce the same model again
  num.threads = parallel::detectCores() - 1) # Use all but one CPU core for quick model training
```
kein regression ausgeben? oder wird regression nur ausgegeben wenn continuous variable habe? und hier habe ja categorical, deshalb dann classification?? habe was falsch geändert??
```{r }
# print a summary of the fitted model
print(rf_basic_wl) 
```
Interpretationsatz einfügen. OBB Wert erwähnen.
with 91 independant variables and splitrule gini an OBB error of 21.49 %.
OBB erklären was ist

### 2.1.2 Evaluation on the test dataset 
Under consideration of appropriate metrics as described in AGDS Book 9.4.1.2.
Das Kapitel als Referenz schreiben, das daraus die Infos habe wie das berechne
confusion matrix machen?
Accuracy testen? 
```{r}
# predictions on the test dataset (basic model, direct class predictions 0/1)
pred_test_basic <- predict(
  rf_basic_wl,                         # basic RF model
  data = df_test_wl[, predictors_all]  # predictors from the test set (all covariates)
)

# extract predicted classes 
pred_class_basic <- pred_test_basic$predictions

# extract true observed classes from the test dataset
obs_class_basic <- df_test_wl[[target_wl]]
```

```{r, fig.cap="Figure 1: Confusion Matrix for the simple model"}
# create confusion matrix
cm_table_basic <- table(
  Observation = factor(obs_class_basic, levels = c(0, 1), labels = c("NoWaterlog", "Waterlog")),
  Prediction  = factor(pred_class_basic, levels = c(0, 1), labels = c("NoWaterlog", "Waterlog"))
)

mosaicplot(cm_table_basic,
  main = "Confusion matrix"
)
#    0  1
# 0 TN  FP
# 1 FN  TP
```

```{r}
# extract True Positives (TP), True Negatives (TN), False Positives (FP), and False Negatives (FN) from the confusion matrix
TP <- cm_table_basic["Waterlog", "Waterlog"]
TN <- cm_table_basic["NoWaterlog", "NoWaterlog"]
FP <- cm_table_basic["NoWaterlog", "Waterlog"]
FN <- cm_table_basic["Waterlog", "NoWaterlog"]

# accuracy
accuracy_basic <- (TP + TN) / (TP + TN + FP + FN)

# precision
precision_basic <- TP / (TP + FP)

# true positive rate (=sensitivity)
tpr_basic <- TP / (TP + FN)

# false positive rate
fpr_basic <- FP / (FP + TN)

# calculated metrics in table format
metrics_basic <- data.frame(
  Metric = c("Accuracy", "Precision", "True positive rate", "False positive rate"),
  Value = c(round(accuracy_basic, 6),
            round(precision_basic, 6),
            round(tpr_basic, 6),
            round(fpr_basic, 6))
)

metrics_basic
```

Analyse der Klassifikationsmetriken bei deinen Daten
1) Prüfen, ob die Daten balanced sind

Balanced bedeutet, dass ungefähr gleich viele Beobachtungen für jede Klasse vorhanden sind:

- `TRUE` = Waterlog  
- `FALSE` = NoWaterlog  

Mit folgendem R-Code kannst du prüfen, ob die Daten balanced sind:
- Werte nahe 0.5/0.5 → balanced
- Werte weit entfernt → imbalanced

The data is not balanced in terms of observed TRUE (waterlogged = 1) and FALSE (not waterlogged = 0) values. This cannot / can??? directly be seen from the confusion matrix und vom matrix output but can be made visibe by the following two code chunks:

```{r}
# absolute frequencies of each class
table(df_test_wl[[target_wl]])
```
Die absoluten Häufigkeiten von observed waterlogged ist XX mal und von nicht waterlogged is XX mal. 
```{r}
# percentage of each class
prop.table(table(df_test_wl[[target_wl]]))
```
Interpretation of Class Balance - The values:
- Class `0` (NoWaterlog): 0.65 → 65% of the observations  
- Class `1` (Waterlog): 0.35 → 35% of the observations

For the interpretation ot the different metrics this implies that
It's therefore important to keep in mind
Conclusion

- The dataset is **imbalanced**, because one class (NoWaterlog) appears almost twice as often as the other (Waterlog).  
- This implies that **accuracy alone can be misleading**, since a model that always predicts the majority class would already achieve 65% accuracy.  
- Metrics like **precision, recall, F1-score, and ROC/AUC** are more informative for evaluating model performance in this case.


Interpretation of Classification Metrics and Your Results

2) Interpretation of Metrics

- **Accuracy**: This is the proportion of all observations that the model correctly classifies. It gives an overall sense of model performance but can be misleading if the data is imbalanced because the majority class can dominate the score.

- **Precision**: This measures how many of the predicted positives (in your case, `Waterlog`) are actually correct. It reflects the reliability of positive predictions.

- **True Positive Rate (Recall / Sensitivity)**: This is the proportion of actual positives that the model correctly identifies. It shows how well the model captures all real `Waterlog` cases.

- **False Positive Rate**: This is the proportion of actual negatives (`NoWaterlog`) that are incorrectly classified as positives. It reflects the model’s tendency to produce false alarms for the negative class.

**General note:**  
- For balanced datasets, accuracy is usually meaningful and roughly aligned with precision and recall.  
- For imbalanced datasets, precision, recall, and FPR are more informative than accuracy, because a high accuracy could be achieved by simply predicting the majority class.

3)  Interpretation of Your Values

Your metrics:

- Accuracy: 0.765 → The model correctly classifies about 76.5% of all cases, which is clearly better than random guessing.  
- Precision: 0.658 → About 65.8% of the predicted `Waterlog` cases are correct. Roughly one-third of positive predictions are false positives.  
- True Positive Rate (Recall): 0.686 → The model identifies approximately 68.6% of all actual `Waterlog` cases.  
- False Positive Rate: 0.192 → About 19.2% of `NoWaterlog` cases are incorrectly classified as `Waterlog`.

**Summary:**  
- The model performs noticeably better than chance.  
- It captures a majority of actual `Waterlog` cases, but there is still room to reduce false positives.  
- Precision being lower than recall suggests the model is slightly overpredicting `Waterlog` compared to the actual occurrences.

## 2.2 Variable selection 
### 2.2.1 Variable importance
--> jetzt zuerst variable importance machen: style = permutation; danach Boruta für selection
Interpretation dazuschreiben, welche variablen die wichtigen sind, also welche ich dann schlussendlich behalten habe -> variablenliste anschauen, die im Tutorial hinterlegt ist
```{r }
# ranger() crashes when using tibbles, so we are using the
# base R notation to enter the data
rf_basic_wl_ <- ranger::ranger( 
  y = df_train_wl[, target_wl],       # target variable
  x = df_train_wl[, predictors_all],  # Predictor variables
  importance   = "permutation",       # with variable importance
  seed = 42,                          # set seed for reproducibility of the same model again
  num.threads = parallel::detectCores() - 1) # use all but one CPU core for quick model training
```

```{r }
# extract the variable importance and create a long tibble
vi_rf_basic_wl <- rf_basic_wl_$variable.importance |>
  dplyr::bind_rows() |> 
  tidyr::pivot_longer(cols = dplyr::everything(), names_to = "variable")
```
stimmt das, dass es variable importance based on OBB ist?? oder based on was anderem?
```{r, fig.height=18, fig.cap="Figure 2: Variable importance based on OBB" }
# plot variable importance, ordered by decreasing value
gg_wl <- vi_rf_basic_wl |> 
  ggplot2::ggplot(ggplot2::aes(x = reorder(variable, value), y = value)) +
  ggplot2::geom_bar(stat = "identity", fill = "grey50", width = 0.75) + 
  ggplot2::labs(
    y = "Change in OOB error after permutation", 
    x = "",
    title = "Variable importance based on OOB") +
  ggplot2::theme_classic() +
  ggplot2::coord_flip()
gg_wl # display plot
```
dazuschreiben was man sieht, also dass welche variablen die wichtigsten sind -> das hier ist variable importance, danach kommt dann boruta selection

### 2.2.2 Boruta variable selection
```{r }
# boruta variable selection
set.seed(42)

# run the algorithm
bor_wl <- Boruta::Boruta(
  y = df_train_wl[, target_wl], 
  x = df_train_wl[, predictors_all],
  maxRuns = 50, # number of iterations, can be set lower for faster computation
  num.threads = parallel::detectCores()-1)
```

```{r }
# obtain results -> data frame with all variables, ordered by their importance
df_bor_wl <- Boruta::attStats(bor_wl) |> 
  tibble::rownames_to_column() |> 
  dplyr::arrange(dplyr::desc(meanImp))
```

```{r, fig.height=18, fig.cap="Figure 3: Boruta variable selection"}
# plot the importance result  
ggplot2::ggplot(ggplot2::aes(x = reorder(rowname, meanImp), 
                             y = meanImp,
                             fill = decision), 
                data = df_bor_wl) +
  ggplot2::geom_bar(stat = "identity", width = 0.75) + 
  ggplot2::scale_fill_manual(values = c("grey30", "tomato", "grey70")) + 
  ggplot2::labs(
    y = "Variable importance", 
    x = "",
    title = "Variable importance based on Boruta") +
  ggplot2::theme_classic() +
  ggplot2::coord_flip()
```

```{r }
# get retained important variables
predictors_selected_wl <- df_bor_wl |> 
  dplyr::filter(decision == "Confirmed") |>
  dplyr::pull(rowname)

length(predictors_selected_wl)
```
--> 39 variablen wurden selected, das sind die rot markierten in dem plot obendrüber--> Interpretationstext schreiben welche Variablen ich nehme und wieso (siehe Variablenliste AGDS Buch, was diese bedeuten). von variable importance von davor wurden jetzt welche ausgewählt anhand von Boruta

```{r }
# retrain random forest model with the 39 selected predictors
rf_bor_wl <- ranger::ranger( 
  y = df_train_wl[, target_wl],              # target variable
  x = df_train_wl[, predictors_selected_wl], # predictor variables
  seed = 42,                                 # set seed for reproducibility of the model
  num.threads = parallel::detectCores() - 1)
```

```{r }
# quick report and performance of retrained model object
print(rf_bor_wl)
```
Ein kleinerer OOB-Fehler ist besser.
Das bedeutet: 20,00 % ist besser als 21,49 % → dein Random-Forest-Modell hat sich verbessert.

Der Out-of-Bag (OOB) Prediction Error ist eine interne Schätzung der Fehlklassifikationsrate des Random Forest.

Ein hoher OOB-Fehler = Modell liegt häufiger falsch → schlechter

Ein niedriger OOB-Fehler = Modell liegt seltener falsch → besser

### 2.2.3 Model performance on test set
compare the model performance on the test set with what was obtained with the model using all available covariates. Which model generalises better to unseen data?

```{r } 
# predictions on the test dataset (retrained model, direct class predictions 0/1)
pred_test_bor <- predict(
  rf_bor_wl,                                  # retrained random forest model
  data = df_test_wl[, predictors_selected_wl] # predictors from the test set
)

# extract predicted classes
pred_class_bor <- pred_test_bor$predictions

# extract true observed classes from the test dataset ---
obs_class_bor <- df_test_wl[[target_wl]]

# create confusion matrix
cm_table_bor <- table(
  Observation = factor(obs_class_bor, levels = c(0, 1), labels = c("NoWaterlog", "Waterlog")),
  Prediction  = factor(pred_class_bor, levels = c(0, 1), labels = c("NoWaterlog", "Waterlog"))
)
```

```{r, fig.cap="Figure 4: Confusion Matrix for Boruta variable selection"} 
mosaicplot(cm_table_bor,
  main = "Confusion matrix"
)
#    0  1
# 0 TN  FP
# 1 FN  TP
```

Interpretation: vergleichen mit figure 1 ob boruta variable selection auch was anzeigt, vielleicht sogar besser? wieso kann ein model mit weniger variablen besser sein? das basic model hatte alle parameter?

```{r }
# extract TP, TN, FP, FN from the confusion matrix
TP <- cm_table_bor["Waterlog", "Waterlog"]
TN <- cm_table_bor["NoWaterlog", "NoWaterlog"]
FP <- cm_table_bor["NoWaterlog", "Waterlog"]
FN <- cm_table_bor["Waterlog", "NoWaterlog"]

# accuracy
accuracy_bor <- (TP + TN) / (TP + TN + FP + FN)

# precision
precision_bor <- TP / (TP + FP)

# true positive rate (=sensitivity)
tpr_bor <- TP / (TP + FN)

# false positive rate
fpr_bor <- FP / (FP + TN)

# calculated metrics in table format
metrics_bor <- data.frame(
  Metric = c("Accuracy", "Precision", "True positive rate", "False positive rate"),
  Value = c(round(accuracy_bor, 6),
            round(precision_bor, 6),
            round(tpr_bor, 6),
            round(fpr_bor, 6))
)

metrics_bor
```

```{r } 
# comparison of OBB prediction errors
rf_basic_wl$prediction.error
rf_bor_wl$prediction.error
```

### 2.2.4 Comparison simple model vs. model with variable selection
**Which model generalises better to unseen data?**
**performance metrics:** confusion matrixes beider outputs, simple model und variable selection anschauen und diskutieren. Auf fig x und fig y verweisen.
schreiben ob false positive rate etc. diese dinge besser sind, also model weniger falsches vorhersagt
Compared to the "simple model": The Boruta model performs better on every single metric, meaning it:
misses fewer true positives,
makes fewer false alarms,
is more precise,
and has higher overall accuracy.
-> This is exactly what you want from a model evaluated on unseen data. The Boruta model clearly generalises better to the test set.
Defintion was OBB ist und was zeigt. how many false 
**OBB comparison:**Hier schreiben, dass geringer OBB wert am besten ist. zahl oben automatischer output mit print(model_name) ergibt in %, hier im codechunk oberhalb nochmal ausgegeben als dezimalzahl (nicht gerundet). Zeigt accuracy bzw. 1-accuracy? mit OBB würde man das gleiche sagen, für boruta ist bisschen besser (also geringerer Wert) als für simple model
Based on the independent test set, the Boruta‐selected model generalises better to unseen data: it achieves higher accuracy, precision, and true positive rate, as well as a lower false positive rate compared to the full model. The out-of-bag (OOB) errors provided by the Random Forest training procedure point in the same direction, with the Boruta model showing a slightly lower OOB error than the full model. While a smaller OOB error generally indicates better predictive performance, relying exclusively on OOB error can be problematic because it is calculated during training and may favour more complex models that do not truly generalise well. In this case, however, both the independent test set and the OOB error consistently indicate that the Boruta model performs slightly better, making it the more reliable choice.
---> frage: 
Would the same model choice be made if we considered the OOB prediction error reported as part of the trained model object? -> nein, dann gibt es mir ein anderes ergebnis aus. aber wieso? wird OOB hier anders berechnet als oben? wieso? basierend auf anderen daten, hier auf testdaten? -> natürlich wird oob dann schlechter, weil ja weniger variablen habe die das beschreiben, also weniger detailliert??

# 3. Model optimization
Using the training data subset, implement a 5-fold cross-validation to optimise the hyperparameters mtry and min.node.size of the same Random Forest model as implemented above. You may use the {caret} library
Evaluate the optimized model on the test set using the same metrics as considered above. Does the model generalise better to unseen data than the initial model (which used default hyperparameters, see ?ranger::ranger) 

For the hyperparameter tuning I use the approach, given in the AGDS book with the common default values mtry = sqrt(p) for classification (p=number of predictions). The values 1,3,5,10,15,20 are used for .min.node.size to avoid possible overfitting when the model creates overly deep trees with very small terminal nodes.

habe xy ausprobiert -> 2x anzeigen
## 3.1 Hyperparameter tuning
```{r }
# define hyperparameter grid for tuning
pp <- recipes::recipe(
  stats::as.formula(
    paste(target_wl, "~", paste(predictors_selected_wl, collapse = "+"))),
                      data = df_train_wl)|>
  recipes::step_center(recipes::all_numeric(), -recipes::all_outcomes()) |>
  recipes::step_scale(recipes::all_numeric(), -recipes::all_outcomes())

set.seed(42)  # set seed for reproducibility

p <- floor(sqrt(length(predictors_selected_wl))) # estimate good mtry values from my data 

tune_grid <- expand.grid(
  .mtry = c(4, p, 8),                   # try different mtry values
  .min.node.size = c(1,3,5,8,10,15,20), # try different min.node.size values 
  .splitrule = "gini"                   # standard splitrule for classification
)
```
sqrt und floor war im AGDS book -> also quelle dazu schreiben wie ich p berechnet habe, in welchem kapitel stand das?
```{r, results='hide'}
# train a random forest with 5-fold cross validation
mod_cv <- caret::train(
  pp,
  data = df_train_wl |> drop_na(),
  method = "ranger",
  trControl = trainControl(method = "cv", number = 5, savePredictions = "final"),
  tuneGrid = tune_grid,
  metric = "Accuracy",
  num.trees = 2000, # may take some computation time, reduce for faster computation
  seed = 42
)
```
ist nicht sensitiv, ist ein wert einfach
```{r }
print(mod_cv) # print cross validation results
```

```{r, fig.cap="Figure 5: Visualisation of the tuning hyperparameters"}
ggplot(mod_cv) # visualize tuning hyperparameters results
```
For the classification of the tuning parameters the splitrule "gini" and the metric "Accuracy" was used. The final values used for the model with a number of 2000 trees, were mtry = 6, and min.node.size = 20, giving an accuracy of 0.7884246. Since the plot does not show a clear separation between the different numbers of randomly selected predictors (the .mtry values), we can assume that the model is not particularly sensitive to the choice of randomly selected predictors.

## 3.2 Evaluation on the test dataset
```{r }
# predictions on the test dataset
pred_class_cv <- predict(mod_cv, newdata = df_test_wl)  

# make sure the variable is a factor and with correct levels (0/1)
pred_class_cv <- factor(pred_class_cv, levels = c(0, 1), labels = c("NoWaterlog", "Waterlog"))

# apply to test dataset classes
true_class_cv <- factor(df_test_wl[[target_wl]],
                        levels = c(0, 1),
                        labels = c("NoWaterlog", "Waterlog"))
```

evaluate model noch: so war im buch
eval_model(mod = mod, df_train = ddf_train, df_test = ddf_test)
oder habe ich mit confusion matrix evaluation gemacht??

```{r, fig.cap="Figure 6: Confusion matrix for optimized RF"}
# confusion matrix
cm_table_cv <- table(
  Observation = true_class_cv,
  Prediction  = pred_class_cv
)

# mosaicplot
mosaicplot(cm_table_cv, main = "Confusion matrix (Optimized RF with tuned hyperparameters)")
```

falls mosaic plot mache, dann auch abbildungsunterschrift mosaicplot for optimised model und muss vergleichen mit den plots von before (boruta und basic am anfang)

```{r }
# evaluation of the optimized model
# TP, TN, FP, FN
TP_cv <- cm_table_cv["Waterlog", "Waterlog"]
TN_cv <- cm_table_cv["NoWaterlog", "NoWaterlog"]
FP_cv <- cm_table_cv["NoWaterlog", "Waterlog"]
FN_cv <- cm_table_cv["Waterlog", "NoWaterlog"]

# accuracy, precision, true positive rate, false positve rate
accuracy_cv <- (TP_cv + TN_cv) / (TP_cv + TN_cv + FP_cv + FN_cv)
precision_cv <- TP_cv / (TP_cv + FP_cv)
tpr_cv <- TP_cv / (TP_cv + FN_cv)
fpr_cv <- FP_cv / (FP_cv + TN_cv)

# show result as table
metrics_cv <- data.frame(
  Metric = c("Accuracy", "Precision", "True positive rate", "False positive rate"),
  Value = c(round(accuracy_cv, 6),
            round(precision_cv, 6),
            round(tpr_cv, 6),
            round(fpr_cv, 6))
)

metrics_cv
```
## 3.3 Comparison initial model vs. model with tuned hyperparameters 
Metrics-ergebnisse hier sind noch besser als die Ergebnisse von davor von boruta und von Anfang simple model
Result Hyperparameters auch ausgeben noch, also welche kamen jetzt raus als gute hyperparameter am ende und welche waren default im model zuerst??

vergleichen mit dem initial model. hier interpretation schreiben wie diese beiden models
also vergleichen mit boruta oder mit simple model? oder mit beiden vergleichen am besten, vermutlich ist eher boruta gemeint, weil mache hier ja auch model optimization auf basis von boruta model. simple model verwende ich ja gar nicht weiter. ja benutze predictors selected dafür, also meine 39 selected.
Does the model generalise better to unseen data than the initial model (which used default hyperparameters, see ?ranger::ranger --> das noch nachschauen was das ausgibt 
stand auch im buch das für klassifikation default mäßig folgendes benutzt wird: 
From the R help documentation one can read that for ranger::ranger the default hyperparameters for classification are as follows: 

# 4. Probabilistic predictions
## 4.1 Random forest model with optimised hyperparameters
```{r}
# train random forest with probability output
# using optimized hyperparameters from mod_cv
rf_bor_wl_prob <- ranger(
  y = df_train_wl[, target_wl],           # target variable
  x = df_train_wl[, predictors_all],      # predictors
  mtry = mod_cv$bestTune$mtry,
  min.node.size = mod_cv$bestTune$min.node.size,
  splitrule = "gini",                     
  num.trees = 500,                       
  probability = TRUE,                    
  seed = 42,
  num.threads = parallel::detectCores() - 1
)
```

!!!!!hierfür brauche quelle wie ich den code wusste, weil steht nicht im AGDS buch, doch in buch für geodata 1, Kapitel angeben als referenz?????????
```{r}
# predict probabilities on the test dataset
pred_prob <- predict(rf_bor_wl_prob, data = df_test_wl)$predictions

# extract probability for class = 1 = soil is waterlogged)
prob_waterlog <- pred_prob[, "1"]

# true class labels as numeric vectors (1=waterlogged soil, 0=no waterlogged soil)
true_class <- as.numeric(as.character(df_test_wl[[target_wl]]))
```

## 4.2 Reicever-operating-characteristic (ROC) curve
I used the information from the pROC package (https://search.r-project.org/CRAN/refmans/pROC/html/pROC-package.html) for plotting the ROC curve.
```{r, results='hide'}
# calculate ROC (receiver operating characteristic curce) & AUC (area under curve)
roc_obj <- roc(true_class, prob_waterlog)
auc_val <- auc(roc_obj)

# prepare for plotting
roc_df <- data.frame(
  FPR = 1 - roc_obj$specificities, # false positive rate
  TPR = roc_obj$sensitivities
)
```

```{r, fig.cap="Figure 7: ROC curve"}
# plot ROC curve
ggplot(roc_df, aes(x = FPR, y = TPR)) +
  geom_line(color = "blue", size = 1.2) +                                      # ROC line
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +  # random classifier line
  labs(
    title = paste0("ROC Curve (AUC = ", round(auc_val, 3), ")"),
    x = "False Positive Rate",
    y = "True Positive Rate"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid.major = element_line(color = "grey80"),
    panel.grid.minor = element_blank()
  )
```

The ROC analysis resulted in an AUC of approximately 0.80, indicating good discriminatory performance of the model.
An AUC value of 0.80 suggests that the model is able to reliably distinguish between waterlogged and non-waterlogged conditions.
-> AUC kann man aber auch weglassen

geom_line() → zeigt die ROC-Kurve (TPR vs. FPR).
geom_abline() → gestrichelte Diagonale, die zufällige Klassifikation darstellt.
AUC wird im Titel angezeigt, gibt die Gesamtleistung an (0.5 = Zufall, 1 = perfekte Trennung).
Je näher die blaue Linie an der oberen linken Ecke, desto besser trennt das Modell Waterlog von NoWaterlog
1. ROC-Kurve allgemein
Die blaue Linie zeigt, wie gut dein Modell zwischen „Waterlog“ und „NoWaterlog“ trennt, wenn du den Schwellenwert von 0 bis 1 variierst.
Die diagonale rote Linie entspricht Zufall (TPR = FPR, also ein vollkommen nutzloses Modell).
Je weiter die blaue Linie links oben liegt, desto besser das Modell.
AUC (Area Under the Curve):
Werte zwischen 0.5 und 1.0.
AUC = 0.5 → Zufall.
AUC = 1 → perfektes Modell.
Je näher an 1, desto besser.
2. Punkte für bestimmte Thresholds
Jeder Punkt zeigt, welche TPR und FPR du bekommst, wenn du genau diesen Threshold verwendest.
Beispiel: Threshold = 0.5
TPR = Anteil der echten Waterlog-Fälle, die korrekt erkannt werden.
FPR = Anteil der NoWaterlog-Fälle, die fälschlicherweise als Waterlog klassifiziert werden.
Niedrigere Thresholds → mehr „positiv“ klassifiziert → höhere TPR, aber auch höhere FPR.
Höhere Thresholds → vorsichtiger → weniger False Positives, aber TPR sinkt.
3. Entscheidungsfindung
Du wählst den Threshold je nach Risiko/Anwendung:
Kritische Anwendung (z.B. Bauprojekt): Lieber niedriger Threshold → keine nassen Böden übersehen → mehr False Positives akzeptieren.
Unkritische Anwendung: Höherer Threshold → weniger Fehlalarme, TPR könnte etwas sinken.
4. Praktische Hinweise
Die Punkte auf der Kurve helfen dir zu visualisieren:
„Wenn ich Threshold = 0.2 nehme, wie viele True Positives vs. False Positives bekomme ich?“
„Wenn ich Threshold = 0.8 nehme, wie konservativ bin ich?“
Du siehst direkt den Trade-off zwischen Sensitivity und Specificity.
Wenn du willst, kann ich dir eine kurze Anleitung schreiben, wie man automatisch den optimalen Threshold nach Youden-Index aus der ROC-Kurve ermittelt und den Punkt auf dem Plot markiert – das wird im AGDS Buch so empfohlen.
Willst du, dass ich das mache?
Ja, genau! Die Threshold-Punkte liegen direkt auf der blauen ROC-Linie.
Das bedeutet:
Jeder Punkt zeigt, welches TPR/FPR-Verhältnis dein Modell erzielt, wenn du genau diesen Threshold wählst.
Die blaue Linie selbst zeigt alle möglichen Thresholds von 0 bis 1.
Die Punkte sind nur Markierungen für bestimmte Schwellenwerte, damit du sehen kannst, wie sich die Klassifikation bei diesen Thresholds verhält.
Was du daraus lesen kannst:
Trade-off zwischen Sensitivity und Specificity
Niedriger Threshold → Punkt weiter oben rechts: hohe Sensitivity (mehr echte Positives erkannt), aber auch hohe FPR (mehr falsche Positives).
Hoher Threshold → Punkt weiter unten links: niedrige FPR (weniger falsche Positives), aber auch niedrige Sensitivity (mehr echte Positives verpasst).
Vergleich verschiedener Thresholds
Die Punkte machen sichtbar, welche Thresholds „aggressiv“ oder „konservativ“ klassifizieren.
Beispiel:
Threshold 0.2 → konservativ für positives Vorhersagen, Punkt weit oben rechts.
Threshold 0.8 → sehr vorsichtig, Punkt weit unten links
Optimale Threshold-Auswahl
Du kannst anhand der Punkte sehen, welcher Threshold einen guten Kompromiss zwischen TPR und FPR bietet.
Oft wird Youden-Index genutzt: Threshold, bei dem (TPR - FPR) maximal ist.
Also kurz gesagt:
Punkte auf der Linie = Modellleistung bei diesen Schwellenwerten.
Höhe links oben = gute Trennung zwischen Klassen.
Bewegung entlang der Linie zeigt, wie sich Sensitivity und FPR ändern, wenn du den Threshold änderst.

TPR (Sensitivity)
1 ┤       •  Threshold 0.2 (aggressiv, viele True Positives, aber auch FPs)
  │      /
  │     /
0.8 ┤    •  Threshold 0.5 (ausgewogen)
  │   /
  │  /
0.6 ┤ •  Threshold 0.8 (vorsichtig, wenige FPs, aber auch weniger TPs)
  │/
0 └──────────────────────── FPR (1 - Specificity)
  0   0.2   0.4   0.6   0.8   1
Blaue Linie = ROC-Kurve für alle möglichen Thresholds

Punkte = bestimmte Thresholds, die du z.B. 0.2, 0.5, 0.8 ausgewählt hast
FPR = Anteil der negativen Fälle, die fälschlicherweise als positiv vorhergesagt wurden
TPR = Anteil der positiven Fälle, die korrekt erkannt wurden
ROC-Kurve zeigt Performance des probabilistischen Modells
-> Interpreation der Kurve: 

threshold interpretation: 
When informing a project where waterlogged soils are critical for stability, a low threshold should be chosen. This prioritizes sensitivity, ensuring most waterlogged areas are detected, even if false alarms occur. In contrast, for a project where waterlogging is unwanted but not critical, a higher threshold is preferable, reducing false positives at the cost of missing some waterlogged areas.

Thresholds should therefore differ depending on the scenario. A similar analogy can be found in medicine: for a critical diagnosis like cancer, a low threshold is used to avoid missing patients, whereas for routine checks of minor conditions, a higher threshold avoids unnecessary interventions.

## 4.3 Choice / Interpretation of threshold
  Consider you inform an infrastructure construction project where waterlogged soils     severely jeopardize the stability of the building. Then, consider you inform a project where waterlogged soils are unwanted, but not critical. In both cases, your prediction map of a binary classification is used as a basis and the binary classification is derived from the probabilistic prediction. How would you chose the threshold in each case? Would you chose the same threshold in both cases? If not, explain why. Can you think of an analogy of a similarly-natured problem in another realm?
- Diskutiere zwei Szenarien:
    1. Kritisch: Wasserlog ist sehr problematisch für ein Bauprojekt → lieber konservativ
    niedriger Schwellenwert, damit keine nassen Böden übersehen werden.
    2. Unkritisch: Wasserlog ist unerwünscht, aber nicht gefährlich → höherer 
    Schwellenwert, um Fehlalarme zu vermeiden.
- Beantworte:
    - Würdest du in beiden Fällen denselben Schwellenwert wählen? (→ Nein, je nach Risiko unterschiedlich.)
    - Gib ein analoges Beispiel aus einem anderen Bereich (z. B. Medizin: Krebsdiagnose vs. Routinekontrolle).
Variable list noch anschauen, die im Tutorial hinterlegt ist

# 5. References & Declaration of Authorship
I hereby declare, that I wrote this report myself. It is based on foundational codes and general information from the tutorial "Digital Soil Mapping - Predicting soil properties from spatial data" by Pascal Schneider, Benjamin Stocker, Madlene Nussbaum (2023), general information from the book "Applied Geodata Science (v1.0)" by Benjamin Stocker et. al (2023) and informations from the lectures (held by Prof. Dr. Benjamin Stocker & Dr. Fabian Bernhard). All additional sources of information can be found in this section and were cited in the text. Last but not least "R Markdown: The Definitive Guide" by Yihui Xie, J. J. Allaire and Garrett Grolemund (2023) (available here: https://bookdown.org/yihui/rmarkdown/r-code.html) was used as a helpful guide on markdown syntax.
?????????? chat für problem with the curve???

**Data**
The Report is based on following original data. Damit preprocessing von data gemacht und dann in ordner abgelegt.
soil data: soil properties (including ?? wieviele variabeln) from locations across canton of Bern. https://github.com/geco-bern/tutorial_digital_soil_mapping
climate: climatic variables (???). Numerisch

**Other Literature**
