---
title: "Report Exercise: Digital Soil Mapping"
author: "Salome Hayler"
date: "22.12.2025"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
options(
  repos = c(CRAN="https://cloud.r-project.org")
)
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
rm(list = ls()) # clear environment
```

# 1. Introduction
Soil conditions vary strongly across space and influence land use and infrastructure planning. One important soil property is waterlogging, which affects e.g. soil stability and drainage. In this exercise, the digital soil mapping workflow, from the tutorial "Digital Soil Mapping - Predicting soil properties from spatial data" by Pascal Schneider, Benjamin Stocker, Madlene Nussbaum (2023) (available here: https://geco-bern.github.io/tutorial_digital_soil_mapping/), using random forests is applied to the binary categorical variable waterlog.100. This variable indicates whether the soil is waterlogged at a depth of 100 cm (TRUE or FALSE) across the area of the canton of Bern. The analysis is based on soil data as point observations of soil properties from field sampling locations combined with a large set of spatially continuous environmental covariates. First, a simple random forest model using all predictors is trained and evaluated on a test dataset. Next, variable selection and hyperparameter optimization using cross validation are applied to improve model performance. Finally, probabilistic predictions are used to explore how different classification thresholds affect decision-making in different application contexts.

# 2. Re-implemention of the digital soil mapping workflow for waterlog.100
The data required for this exercise was downloaded from the tutorial_digital_soil_mapping and preprocessed beforehand following the instructions of the tutorial and then stored in the data folder. The R Markdown file is linked to this folder, while the full code for preprocessing the raw data (see data_raw) is not shown in the rendered document.

```{r}
# required packages
packages <- c(
  "dplyr",
  "tidyr",
  "readr",
  "tibble",
  "terra",
  "sf",
  "tidyterra",
  "ggplot2",
  "lattice",
  "ranger",
  "Boruta",
  "caret",
  "recipes",
  "pROC",
  "knitr",
  "rmarkdown",
  "here"
)
```

```{r, include=FALSE}
# install missing packages
for (pkg in packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
}

# load frequently used packages
library(dplyr)      # for data manipultations
library(tidyr)      # for data transformations
library(terra)      # for processing raster data
library(here)       # for reproducibility
library(ggplot2)    # for making plots
library(caret)      # for cross-validation, hyperparameter tuning
library(ranger)     # for random forest
library(Boruta)     # for variable selection
library(sf)         # for spatial features
library(pROC)       # for ROC/AUC
library(knitr)      # for rendering R Markdown
library(patchwork)  # for combining multiple plots
library(lattice)    # for intern plots e.g. during cross validation

# set project root
here::i_am("agds_report_salomehayler.Rproj")
```

```{r, eval=FALSE, include=FALSE}
# load soil data from sampling locations
df_obs <- readr::read_csv(
  here::here("data_raw/soildata/berne_soil_sampling_locations.csv")
  )

# display data
head(df_obs) |> 
  knitr::kable()
```

```{r, eval=FALSE, include=FALSE}
# load the covariates that we want to produce our soil maps with. These files are in the geoTIFF format - geolocated TIFF files.
# get a list with the path to all raster files
list_raster <- list.files(
  here::here("data_raw/geodata/covariates/"),
  full.names = TRUE
  )

# display data (lapply to clean names)
lapply(
  list_raster, 
  function(x) sub(".*/(.*)", "\\1", x)
  ) |> 
  unlist() |> 
  head(5) |> 
  print()
```

```{r, eval=FALSE, include=FALSE}
# load a raster file as example: Picking the slope profile at 2 m resolution
raster_example <- terra::rast(
  here::here("data_raw/geodata/covariates/Se_slope2m.tif")
  )
raster_example
```

```{r , eval=FALSE, include=FALSE}
# combine data: The {terra} package comes with the very useful tool to stack multiple rasters on top of each other if they share the spatial grid (extent and resolution). To do so, we just have to feed in the vector of file names list_raster
# load all files as one batch
all_rasters <- terra::rast(list_raster)
all_rasters
```

```{r , eval=FALSE, include=FALSE}
# extract coordinates from sampling locations
sampling_xy <- df_obs |> 
  dplyr::select(x, y)

# from all rasters, extract values for sampling coordinates
df_covars <- terra::extract(
  all_rasters,  # the raster we want to extract from
  sampling_xy,  # a matrix of x and y values to extract for
  ID = FALSE    # to not add a default ID column to the output
  )

df_full <- cbind(df_obs, df_covars)
head(df_full) |> 
  knitr::kable() 
```

```{r , eval=FALSE, include=FALSE}
vars_categorical <- df_covars |> 
  
  # get number of distinct values per variable
  dplyr::summarise(dplyr::across(dplyr::everything(), ~dplyr::n_distinct(.))) |> 
  
  # turn df into long format for easy filtering
  tidyr::pivot_longer(
    dplyr::everything(), 
    names_to = "variable", 
    values_to = "n"
    ) |> 
  
  # tilter out variables with 10 or less distinct values
  dplyr::filter(n <= 10) |>
  
  # extract the names of these variables
  dplyr::pull('variable')

cat("Variables with less than 10 distinct values:", 
    ifelse(length(vars_categorical) == 0, "none", vars_categorical))
```

```{r , eval=FALSE, include=FALSE}
# now that we have the names of the categorical values, we can mutate these columns in our data frame using the base function as.factor():
df_full <- df_full |> 
  dplyr::mutate(dplyr::across(all_of(vars_categorical), ~as.factor(.)))
```

```{r, eval=FALSE, include=FALSE}
# save the data
if (!dir.exists(here::here("data"))) system(paste0("mkdir ", here::here("data")))
saveRDS(df_full, 
        here::here("data/df_full.rds"))
```

## 2.1 Simple model for waterlog.100
waterlog.100 is a binary categorical variable that indicates whether the soil is waterlogged at a depth of 100 cm. In this workflow, it is encoded as 0 (not waterlogged) and 1 (waterlogged).

### 2.1.1 Random forest for waterlog.100
```{r, results='hide'}
# load dataframe
df_full_soil <- readRDS(here::here("data/df_full.rds"))

df_full_soil |> 
  knitr::kable()
```

```{r }
# specify target: binary waterlog.100
target_wl <- "waterlog.100"
```

```{r}
# make as factor
df_full_soil[[target_wl]] <- factor(df_full_soil[[target_wl]])
```

```{r }
# specify predictors_all: Remove soil sampling and observational data
predictors_all <- names(df_full_soil)[14:ncol(df_full_soil)]

cat("The target is:", target_wl,
    "\nThe predictors_all are:", paste0(predictors_all[1:8], sep = ", "), "...")
```

```{r }
# split dataset into training and testing sets
df_train_wl <- df_full_soil |> dplyr::filter(dataset == "calibration")
df_test_wl  <- df_full_soil |> dplyr::filter(dataset == "validation")
```

```{r }
# encode target variable as factor
df_train_wl[[target_wl]] <- factor(df_train_wl[[target_wl]]) 
df_test_wl[[target_wl]]  <- factor(df_test_wl[[target_wl]]) 
```

```{r }
# filter out any NA to avoid error when running a random forest
df_train_wl <- df_train_wl |> tidyr::drop_na()
df_test_wl  <- df_test_wl   |> tidyr::drop_na()
```

```{r }
# a little bit of verbose output
n_tot <- nrow(df_train_wl) + nrow(df_test_wl)

perc_cal <- (nrow(df_train_wl) / n_tot) |> round(2) * 100
perc_val <- (nrow(df_test_wl)  / n_tot) |> round(2) * 100

cat("For model training, we have a calibration / validation split of: ",
    perc_cal, "/", perc_val, "%")
```

```{r }
# ranger() crashes when using tibbles, so we are using the
# base R notation to enter the data
rf_basic_wl <- ranger::ranger( 
  y = df_train_wl[, target_wl],       # target variable
  x = df_train_wl[, predictors_all],  # Predictor variables
  seed = 42,                          # Specify seed for randomization to reproduce the same model again
  num.threads = parallel::detectCores() - 1) # Use all but one CPU core for quick model training
```

```{r }
# print a summary of the fitted model
print(rf_basic_wl) 
```
The basic random forest model for waterlog.100 used 91 predictor variables and the gini split rule. The out-of-bag (OOB) prediction error is 21.49%, which gives an estimate of how often the model makes wrong predictions on new data. This number means the model misclassified about 1 in 5 samples during training.

### 2.1.2 Model performance on the test dataset 
In the following the model is evaluated on the test data, using appropriate metrics, including the visualisation of a confusion matrix, from AGDS Book Chapter 9.4.1.2. (https://geco-bern.github.io/agds_book/regression_classification.html#metrics-for-classification).
```{r}
# predictions on the test dataset (basic model, direct class predictions 0/1)
pred_test_basic <- predict(
  rf_basic_wl,                         # basic RF model
  data = df_test_wl[, predictors_all]  # predictors from the test set (all covariates)
)

# extract predicted classes 
pred_class_basic <- pred_test_basic$predictions

# extract true observed classes from the test dataset
obs_class_basic <- df_test_wl[[target_wl]]
```

```{r, fig.cap="Figure 1: Confusion Matrix for the simple model."}
# create confusion matrix
cm_table_basic <- table(
  Observation = factor(obs_class_basic, levels = c(0, 1), labels = c("NoWaterlog", "Waterlog")),
  Prediction  = factor(pred_class_basic, levels = c(0, 1), labels = c("NoWaterlog", "Waterlog"))
)

mosaicplot(cm_table_basic,
  main = "Confusion matrix (basic model)"
)
#    0  1
# 0 TN  FP
# 1 FN  TP
```

```{r}
# extract True Positives (TP), True Negatives (TN), False Positives (FP), and False Negatives (FN) from the confusion matrix
TP <- cm_table_basic["Waterlog", "Waterlog"]
TN <- cm_table_basic["NoWaterlog", "NoWaterlog"]
FP <- cm_table_basic["NoWaterlog", "Waterlog"]
FN <- cm_table_basic["Waterlog", "NoWaterlog"]

# accuracy
accuracy_basic <- (TP + TN) / (TP + TN + FP + FN)

# precision
precision_basic <- TP / (TP + FP)

# true positive rate (=sensitivity)
tpr_basic <- TP / (TP + FN)

# false positive rate
fpr_basic <- FP / (FP + TN)

# calculated metrics in table format
metrics_basic <- data.frame(
  Metric = c("Accuracy", "Precision", "True positive rate", "False positive rate"),
  Value = c(round(accuracy_basic, 6),
            round(precision_basic, 6),
            round(tpr_basic, 6),
            round(fpr_basic, 6))
)

metrics_basic
```
The basic model has an **accuracy** (overall proportion of correct predictions) of **76%**, with a **precision** (proportion of predicted waterlogged soils that are actually waterlogged) of **65%**, a **true positive rate** (proportion of actual waterlogged soils correctly predicted) of **69%**, and a **false positive rate** (proportion of non-waterlogged soils incorrectly predicted as waterlogged) of **20%**. This means the model correctly predicts most cases, but some waterlogged soils are still missed. The same information can be seen in Figure 1 (confusion matrix above), where the largest block represents the correctly predicted non-waterlogged cases (TRUE negatives).

```{r}
# absolute frequencies of each class
table(df_test_wl[[target_wl]])
```

```{r}
# percentage of each class
prop.table(table(df_test_wl[[target_wl]]))
```
The absolute frequencies of the observed classes in the test data are 70 for waterlogged (class 1) and 130 for non-waterlogged (class 0) soils. In percentages, this corresponds to 35% waterlogged and 65% non-waterlogged soils in the dataset. This means the data is not balanced because there are more non-waterlogged samples (130) than waterlogged samples (70). A balanced dataset would have roughly the same number of samples in each class. This means that accuracy alone can be misleading, since the model could be mostly correct by predicting the majority class (that is the non-waterlogged soils). Therefore it is important to also look at precision, true positive rate, and false positive rate to see how well the model predicts the waterlogged soils.

Overall, the basic model performs better than chance and finds most of the actual waterlogged soils, but it still predicts some areas as waterlogged that are not. Because the precision is lower than the true positive rate, this means that the model predicts waterlogged soils a little too often compared to the real cases.

## 2.2 Variable selection 
This section follows the tutorial to make the model simpler. First, variable importance is checked using the permutation method. Then Boruta is used to select the most important predictors. The model is retrained and tested to see how it performs compared to the model using all variables. The OOB error is also looked at to check the model choice.

### 2.2.1 Variable importance
Variable importance is first calculated using the permutation method to see which predictors contribute most to the model.
```{r }
# ranger() crashes when using tibbles, so we are using the
# base R notation to enter the data
rf_basic_wl_ <- ranger::ranger( 
  y = df_train_wl[, target_wl],       # target variable
  x = df_train_wl[, predictors_all],  # Predictor variables
  importance   = "permutation",       # with variable importance
  seed = 42,                          # set seed for reproducibility of the same model again
  num.threads = parallel::detectCores() - 1) # use all but one CPU core for quick model training
```

```{r }
# extract the variable importance and create a long tibble
vi_rf_basic_wl <- rf_basic_wl_$variable.importance |>
  dplyr::bind_rows() |> 
  tidyr::pivot_longer(cols = dplyr::everything(), names_to = "variable")
```

```{r, fig.height=18, fig.cap="Figure 2: Variable importance based on OBB." }
# plot variable importance, ordered by decreasing value
gg_wl <- vi_rf_basic_wl |> 
  ggplot2::ggplot(ggplot2::aes(x = reorder(variable, value), y = value)) +
  ggplot2::geom_bar(stat = "identity", fill = "grey50", width = 0.75) + 
  ggplot2::labs(
    y = "Change in OOB error after permutation", 
    x = "",
    title = "Variable importance based on OOB") +
  ggplot2::theme_classic() +
  ggplot2::coord_flip()
gg_wl # display plot
```
Figure 2 shows that NegO (Topographic negative openness) ist the most important variable, followed by mrvbf25 (Multi-resolution valley bottom flatness at 25m resolution), Se_MRVBF2m (Multi-resolution valley bottom flatness at 2 m resolution), be_gwn25_vdist (Horizontal distance to water body at 25m resolution), Se_TWI2m (Topographic wetness index at 2 m resolution), mt_rr_y (Mean annual temperature) etc... 
This means these variables contribute the most to the model’s predictions and explain the most variation in whether a soil is waterlogged or not (most important variables at predicting the target variable waterlog.100). (Explanations for the other variables here: https://geco-bern.github.io/tutorial_digital_soil_mapping/99-variables.html).

### 2.2.2 Boruta variable selection
In the following, Boruta as a variable selection method is used to identify the most important variables and remove predictors that do not significantly improve the model.
```{r }
# boruta variable selection
set.seed(42)

# run the algorithm
bor_wl <- Boruta::Boruta(
  y = df_train_wl[, target_wl], 
  x = df_train_wl[, predictors_all],
  maxRuns = 50, # number of iterations, can be set lower for faster computation
  num.threads = parallel::detectCores()-1)
```

```{r }
# obtain results -> data frame with all variables, ordered by their importance
df_bor_wl <- Boruta::attStats(bor_wl) |> 
  tibble::rownames_to_column() |> 
  dplyr::arrange(dplyr::desc(meanImp))
```

```{r, fig.height=18, fig.cap="Figure 3: Boruta variable selection."}
# plot the importance result  
ggplot2::ggplot(ggplot2::aes(x = reorder(rowname, meanImp), 
                             y = meanImp,
                             fill = decision), 
                data = df_bor_wl) +
  ggplot2::geom_bar(stat = "identity", width = 0.75) + 
  ggplot2::scale_fill_manual(values = c("grey30", "tomato", "grey70")) + 
  ggplot2::labs(
    y = "Variable importance", 
    x = "",
    title = "Variable importance based on Boruta") +
  ggplot2::theme_classic() +
  ggplot2::coord_flip()
```

```{r }
# get retained important variables
predictors_selected_wl <- df_bor_wl |> 
  dplyr::filter(decision == "Confirmed") |>
  dplyr::pull(rowname)

length(predictors_selected_wl)
```
Using boruta, 39 variables were selected from the original set of 91 predictors for predicting the target variable waterlog.100. See figure 3 and descriptions under https://geco-bern.github.io/tutorial_digital_soil_mapping/99-variables.html for description of the abbreviations.

```{r }
# retrain random forest model with the 39 selected predictors
rf_bor_wl <- ranger::ranger( 
  y = df_train_wl[, target_wl],              # target variable
  x = df_train_wl[, predictors_selected_wl], # predictor variables
  seed = 42,                                 # set seed for reproducibility of the model
  num.threads = parallel::detectCores() - 1)
```

```{r }
# quick report and performance of retrained model object
print(rf_bor_wl)
```
The boruta based random forest model for waterlog.100 used 39 predictor variables. The out-of-bag (OOB) prediction error is 20.00% then, which is slightly lower and therefore better than the error of the basic model (which was 21.49%, see chapter 2.1.1).

### 2.2.3 Model performance on the test dataset
Now, the boruta model performance is tested on the test dataset (similar to what has been done in chapter 2.1.2 for the basic model).

```{r } 
# predictions on the test dataset (retrained model, direct class predictions 0/1)
pred_test_bor <- predict(
  rf_bor_wl,                                  # retrained random forest model
  data = df_test_wl[, predictors_selected_wl] # predictors from the test set
)

# extract predicted classes
pred_class_bor <- pred_test_bor$predictions

# extract true observed classes from the test dataset ---
obs_class_bor <- df_test_wl[[target_wl]]

# create confusion matrix
cm_table_bor <- table(
  Observation = factor(obs_class_bor, levels = c(0, 1), labels = c("NoWaterlog", "Waterlog")),
  Prediction  = factor(pred_class_bor, levels = c(0, 1), labels = c("NoWaterlog", "Waterlog"))
)
```

```{r, fig.cap="Figure 4: Confusion Matrix for Boruta variable selection."} 
mosaicplot(cm_table_bor,
  main = "Confusion matrix (boruta based model)"
)
#    0  1
# 0 TN  FP
# 1 FN  TP
```

```{r }
# extract TP, TN, FP, FN from the confusion matrix
TP <- cm_table_bor["Waterlog", "Waterlog"]
TN <- cm_table_bor["NoWaterlog", "NoWaterlog"]
FP <- cm_table_bor["NoWaterlog", "Waterlog"]
FN <- cm_table_bor["Waterlog", "NoWaterlog"]

# accuracy
accuracy_bor <- (TP + TN) / (TP + TN + FP + FN)

# precision
precision_bor <- TP / (TP + FP)

# true positive rate (=sensitivity)
tpr_bor <- TP / (TP + FN)

# false positive rate
fpr_bor <- FP / (FP + TN)

# calculated metrics in table format
metrics_bor <- data.frame(
  Metric = c("Accuracy", "Precision", "True positive rate", "False positive rate"),
  Value = c(round(accuracy_bor, 6),
            round(precision_bor, 6),
            round(tpr_bor, 6),
            round(fpr_bor, 6))
)

metrics_bor
```
The boruta based model has an **accuracy** of **78.5%**, with a precision of **68%**, a **true positive rate** of **72.9%**, and a **false positive rate** of **18.5%**. The same pattern can be seen in Figure 2 (confusion matrix above), where the correctly predicted cases dominate.

### 2.2.4 Comparison: Simple (basic) model vs. model with variable selection (boruta)
**Which model generalises better to unseen data?**

**Performance metrics:** Comparing the performance metrics of the two models shows that the boruta based model performs slightly better on the test data than the basic model. It has higher accuracy (78.5% vs. 76%), higher precision (68% vs. 64.9%), a higher true positive rate (72.9% vs. 68.6%), and a lower false positive rate (18.5% vs. 20%) than the basic model. This shows a slight improvement compared to the basic model, with about 1.5% fewer false predictions. This indicates that the boruta model generalizes slightly better to unseen data. However, the improvement is very small, and when visually comparing the two confusion matrixes (Figure 1 and Figure 4), this difference is not noticeable.

```{r } 
# comparison of OBB prediction errors
rf_basic_wl$prediction.error
rf_bor_wl$prediction.error
```

**OBB comparison:** 
The out-of-bag (OOB) prediction error for the basic model is 0.214876 (~21.49%), while the boruta based model has an OOB error of 0.2 (20.00%). Since the Boruta model has a slightly lower OOB error, so YES, it would still be chosen as the better model if the decision was based only on OOB error. However, the difference is very small, so both models perform quite similarly according to the OBB results.

# 3. Model optimization
Using the training data subset, implement a 5-fold cross-validation to optimise the hyperparameters mtry and min.node.size of the same Random Forest model as implemented above. You may use the {caret} library
Evaluate the optimized model on the test set using the same metrics as considered above. Does the model generalise better to unseen data than the initial model (which used default hyperparameters, see ?ranger::ranger) 

## 3.1 Hyperparameter tuning
For the hyperparameter tuning I use the approach, given in the AGDS book (this chapter: https://geco-bern.github.io/agds_book/randomforest.html#from-trees-to-a-forest) with the common default values mtry = sqrt(p) for classification (p=number of predictions). The values 1,3,5,10,15,20 are used for .min.node.size to avoid possible overfitting when the model creates overly deep trees with very small terminal nodes.

```{r }
# define hyperparameter grid for tuning
pp <- recipes::recipe(
  stats::as.formula(
    paste(target_wl, "~", paste(predictors_selected_wl, collapse = "+"))),
                      data = df_train_wl)|>
  recipes::step_center(recipes::all_numeric(), -recipes::all_outcomes()) |>
  recipes::step_scale(recipes::all_numeric(), -recipes::all_outcomes())

set.seed(42)  # set seed for reproducibility

p <- floor(sqrt(length(predictors_selected_wl))) # estimate good mtry values from my data 

tune_grid <- expand.grid(
  .mtry = c(4, p, 8),                   # try different mtry values
  .min.node.size = c(1,3,5,8,10,15,20), # try different min.node.size values 
  .splitrule = "gini"                   # standard splitrule for classification
)
```

```{r, message=FALSE}
# train a random forest with 5-fold cross validation
set.seed(42)  # set seed for reproducibility

mod_cv <- caret::train(
  pp,
  data = df_train_wl |> drop_na(),
  method = "ranger",
  trControl = trainControl(method = "cv", number = 5, savePredictions = "final"),
  tuneGrid = tune_grid,
  metric = "Accuracy",
  num.trees = 2000 # may take some computation time, reduce for faster computation
)
```

```{r, message=FALSE, warning=FALSE}
print(mod_cv) # print cross validation results
```

```{r, message=FALSE, warning=FALSE, fig.cap="Figure 5: Visualisation of the tuning hyperparameters."}
ggplot(mod_cv) # visualize tuning hyperparameters results
```

For the classification of the tuning parameters the splitrule "gini" and the metric "Accuracy" was used. The final values used for the model with a number of 2000 trees, were mtry = 8, and min.node.size = 5, giving an accuracy of 0.7982676. The plot (figure 5) shows that the differences in cross-validation accuracy between the different mtry-values (4, 6, 8) are very small (varies only slightly between about 0.785 and 0.798). (This indicates that the performance of the model is not very sensitive (not really big differences in accuracy) to the choice of the number of randomly selected predictors.)

## 3.2 Evaluation on the test dataset
```{r }
# predictions on the test dataset
pred_class_cv <- predict(mod_cv, newdata = df_test_wl)  

# make sure the variable is a factor and with correct levels (0/1)
pred_class_cv <- factor(pred_class_cv, levels = c(0, 1), labels = c("NoWaterlog", "Waterlog"))

# apply to test dataset classes
true_class_cv <- factor(df_test_wl[[target_wl]],
                        levels = c(0, 1),
                        labels = c("NoWaterlog", "Waterlog"))
```

```{r, fig.cap="Figure 6: Confusion matrix for optimized random forest model."}
# confusion matrix
cm_table_cv <- table(
  Observation = true_class_cv,
  Prediction  = pred_class_cv
)

# mosaicplot
mosaicplot(cm_table_cv, main = "Confusion matrix (optimized RF with tuned hyperparameters)")
```

```{r }
# evaluation of the optimized model
# TP, TN, FP, FN
TP_cv <- cm_table_cv["Waterlog", "Waterlog"]
TN_cv <- cm_table_cv["NoWaterlog", "NoWaterlog"]
FP_cv <- cm_table_cv["NoWaterlog", "Waterlog"]
FN_cv <- cm_table_cv["Waterlog", "NoWaterlog"]

# accuracy, precision, true positive rate, false positve rate
accuracy_cv <- (TP_cv + TN_cv) / (TP_cv + TN_cv + FP_cv + FN_cv)
precision_cv <- TP_cv / (TP_cv + FP_cv)
tpr_cv <- TP_cv / (TP_cv + FN_cv)
fpr_cv <- FP_cv / (FP_cv + TN_cv)

# show result as table
metrics_cv <- data.frame(
  Metric = c("Accuracy", "Precision", "True positive rate", "False positive rate"),
  Value = c(round(accuracy_cv, 6),
            round(precision_cv, 6),
            round(tpr_cv, 6),
            round(fpr_cv, 6))
)

metrics_cv
```

The optimized model achieved an accuracy of 78.0%, with a precision of 67.1%, a true positive rate of 72.9%, and a false positive rate of 19.2%. A similar pattern is visible in Figure 6 (confusion matrix of the optimised model), where it can be seen that the correctly predicted cases are the majority.

## 3.3 Comparison initial model (basic model) vs. model with tuned hyperparameters 
From the R help documentation, one can read that for ranger::ranger the default hyperparameters for classification are as follows: num.trees = 500, mtry = sqrt(p) (rounded) (where p is the number of predictors), min.node.size = 1 , and splitrule = gini. 
Using these defaults, the initial basic model achieved an accuracy of 76.0%, a precision of 64.9%, a true positive rate of 68.6%, and a false positive rate of 20.0%.
The optimized model, in which hyperparameters (.mtry and min.node.size) were tuned, achieved slightly higher accuracy (78.0%), precision (67.1%), and true positive rate (72.9%), and slightly lower false positive rate (19.2%). This indicates that hyperparameter tuning provided a modest improvement over the basic model. 
Looking at the confusion matrix of the basic model (Figure 1) and the optimized model (Figure 6), the false positives and false negatives are a bit smaller in the optimized model. But the differences are small, so it is hard to see a clear improvement just from the confusion matrices.

*Comment on the performance of the boruta model (no hyperparameter tuning): Using 39 selected features, performed better than the hyperparameter-tuned model based on the same features. It achieved 78.5% accuracy, 68% precision, 72.9% true positive rate, and 18.5% false positive rate, while the tuned model was slightly worse. This indicates that the Boruta model was robust with respect to the default hyperparameters of the random forest, meaning that its predictive performance was already stable and not highly sensitive to changes in mtry or min.node.size. But there are other reasons why hyperparameter tuning is important: it can improve model efficiency and reduce computation time by avoiding overly complex trees and can also (not in my case here but in general) improve the model performance when the default settings are not optimal for the specific dataset (for example class imbalances or too complex data structures).*

# 4. Probabilistic predictions
## 4.1 Random forest model with optimised hyperparameters
```{r}
# train random forest with probability output
# using optimized hyperparameters from mod_cv
rf_bor_wl_prob <- ranger(
  y = df_train_wl[, target_wl],           # target variable
  x = df_train_wl[, predictors_all],      # predictors
  mtry = mod_cv$bestTune$mtry,
  min.node.size = mod_cv$bestTune$min.node.size,
  splitrule = "gini",                     
  num.trees = 500,                       
  probability = TRUE,                    
  seed = 42,
  num.threads = parallel::detectCores() - 1
)
```

```{r}
# predict probabilities on the test dataset
pred_prob <- predict(rf_bor_wl_prob, data = df_test_wl)$predictions

# extract probability for class = 1 = soil is waterlogged)
prob_waterlog <- pred_prob[, "1"]

# true class labels as numeric vectors (1=waterlogged soil, 0=no waterlogged soil)
true_class <- as.numeric(as.character(df_test_wl[[target_wl]]))
```

## 4.2 Reicever-operating-characteristic (ROC) curve
I used the information from the pROC package (https://search.r-project.org/CRAN/refmans/pROC/html/pROC-package.html) as well as information from other papers (see section 5) for plotting and understanding the ROC curve.
```{r, message=FALSE, warning=FALSE}
# calculate ROC (receiver operating characteristic curce) & AUC (area under curve)
roc_obj <- roc(true_class, prob_waterlog)
auc_val <- auc(roc_obj)

# prepare for plotting
roc_df <- data.frame(
  FPR = 1 - roc_obj$specificities, # false positive rate
  TPR = roc_obj$sensitivities
)
```

```{r, message=FALSE, warning=FALSE, fig.cap="Figure 7: ROC curve."}
# plot ROC curve
ggplot(roc_df, aes(x = FPR, y = TPR)) +
  geom_line(color = "blue", linewidth = 1.2) +                                      # ROC line
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +  # random classifier line
  labs(
    title = paste0("ROC Curve (AUC = ", round(auc_val, 3), ")"),
    x = "False Positive Rate",
    y = "True Positive Rate"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid.major = element_line(color = "grey80"),
    panel.grid.minor = element_blank()
  )
```
In the plot, the ROC (Receiver Operating Characteristic) curve shows the true positive rate (TPR) versus the false positive rate (FPR) for different threshold values. The diagonal red line represents a random classifier (AUC = 0.5), while the blue line illustrates the model’s performance. The closer the blue line is to the top-left corner, the better the separation between the two classes waterlogged/not waterlogged (points higher on the line (toward the top-left) mean the model correctly identifies more waterlogged areas (higher TPR) while making fewer mistakes on non-waterlogged areas (lower FPR)).The AUC for my model is approximately 0.8 , indicating a good ability to distinguish between waterlogged and non-waterlogged conditions (AUC of 1 would indicate a perfect model). The blue line also demonstrates how TPR and FPR change across all possible thresholds. The threshold is the probability cut-off used to turn predicted probabilities into a yes/no classification. A lower threshold labels more cases as positive (catches more true positives but also more false positives) and a higher threshold is more conservative (reduces false positives but missing some true positives). 

## 4.3 Choice / Interpretation of threshold
For a project where waterlogged soils critically affect stability, a low threshold should be chosen to maximize the detection of all waterlogged areas, even if it increases the false alarms. For a project where waterlogging is unwanted but not critical, a higher threshold is preferable to reduce false positives, with accepting that some waterlogged areas may be missed. The thresholds are different for eacht case because the risk tolerance differs. 
Another analogy can be found in medical testing: For example for a critical disease like cancer, a low threshold is used to avoid missing cases, while for minor conditions (e.g. a normal mild winter cold) missing a few cases is not dangerous, and it avoids unnecessary interventions. This is similar to choosing a higher threshold for a non-critical waterlogging project.

# 5. References & Declaration of Authorship
I hereby declare, that I (Salome Hayler) did this report by myself. It is based on foundational codes and general information from the tutorial "Digital Soil Mapping - Predicting soil properties from spatial data" by Pascal Schneider, Benjamin Stocker, Madlene Nussbaum (2023) (https://geco-bern.github.io/tutorial_digital_soil_mapping/), general information from the book "Applied Geodata Science (v1.0)" by Benjamin Stocker et. al (2023) and informations from the lectures (held by Prof. Dr. Benjamin Stocker & Dr. Fabian Bernhard). All additional sources of information can be found in this section and were cited in the text. AI (ChatGPT) was consulted for resolving technical difficulties that could not be addressed otherwise, which includes for this report especially the pROC package, that didn't work in the beginning to plot the ROC curve. Last but not least "R Markdown: The Definitive Guide" by Yihui Xie, J. J. Allaire and Garrett Grolemund (2023) (available here: https://bookdown.org/yihui/rmarkdown/) was used as a helpful guide on markdown syntax.

**Data**

The report is based on the following original data from https://github.com/geco-bern/tutorial_digital_soil_mapping (in the folder data_raw), which were preprocessed prior to the further analysis (see code chunks in the beginning of the skript) and then stored in the folder data. 

*Soil data:* Soil observations from sampling locations across the canton of Bern, including the binary variable waterlog.100, which indicates whether soils are waterlogged at 100 cm depth.

*Environmental covariates:* This dataset contains originally 99 spatial environmental covariates provided as raster layers. (Full list on content here: https://geco-bern.github.io/tutorial_digital_soil_mapping/99-variables.html). After preprocessing / cleaning the original data only 91 variables remained that were available for the further use.

**Additional Literature**
In addition to the tutorial "Digital soil mapping - Predicting soil properties form spatial data" and the AGDS book chapters on random forests, model evaluation, variable selection, classification metrics, hyperparameter tuning some other sources were consulted for this report.

The R documentations for the ranger package and the pROC curve were used to get an understanding of the random forest implementation and ROC curves.

Regarding the latter, two additional articles were also helpful: Fawcett, T. (2006). An introduction to ROC analysis. Pattern Recognition Letters, 27(8), 861-874. https://doi.org/10.1016/j.patrec.2005.10.010. And Robin, X. et al. (2011). pROC: an open-source package for R and S+ to analyze and compare ROC curves. BMC Bioinformatics 12(77). https://doi.org/10.1186/1471-2105-12-77.
