---
title: "Exercise_8.2_Phenology_Modelling"
author: "Salome Hayler"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I will use the phenocamr package which interfaces with the phenocam network API
to download time series of vegetation greenness and derived phenology metrics.

```{r load libraries}
library(phenocamr)
library(ggplot2)
library(dplyr)
library(patchwork)
library(keyring)
library(appeears)
library(sf)
library(leaflet)
```

Download greenness time series, calculate phenology (phenophases), amend with DAYMET data

```{r download data}
phenocamr::download_phenocam(
  site = "harvard$",
  veg_type = "DB",
  roi_id = "1000",
  daymet = TRUE,
  phenophase = TRUE,
  trim = 2022,
  out_dir = tempdir()
  )
```

```{r, results='hide'}
harvard_phenocam_data <- readr::read_csv(
  file.path(tempdir(), "harvard_DB_1000_3day.csv"), 
  comment = "#"
  )
```

```{r }
# reading in harvard phenology only retaining
# spring (rising) phenology for the GCC 90th
# percentile time series (the default)
harvard_phenology <- readr::read_csv(
  file.path(
    tempdir(),
    "harvard_DB_1000_3day_transition_dates.csv"
    ),
  comment = "#"
) |>
  dplyr::filter(
    direction == "rising",
    gcc_value == "gcc_90"
  )
```

Plot time series of the 90th percentile 3-daily green chromatic coordinate GCC

```{r plot}
ggplot(harvard_phenocam_data) +
  geom_line(
    aes(
      as.Date(date),
      smooth_gcc_90
    ),
    colour = "grey25"
  ) +
  geom_point(
    data = harvard_phenology,
    aes(
      as.Date(transition_25),
      threshold_25
    )
  ) +
  labs(
    x = "",
    y = "GCC"
  ) +
  theme_bw() +
  theme(
    legend.position = "none"
  )
```
### Preprocess data

```{r }
# growing degree days = cumulative sum of temperatures above a specified threshold, mostly T0=5°C
# return mean daily temperature as well
# as formal dates (for plotting)
harvard_temp <- harvard_phenocam_data |>
  group_by(year) |>
  dplyr::mutate(
    tmean = (tmax..deg.c. + tmin..deg.c.)/2
  ) |> 
  dplyr::mutate(
    date = as.Date(date),
    gdd = cumsum(ifelse(tmean >= 5, tmean - 5, 0))
  ) |>
  dplyr::select(
    date,
    year,
    tmean,
    gdd
  ) |>
  ungroup()
```

```{r }
# convert the harvard phenology data and only
# retain required data
harvard_phenology <- harvard_phenology |>
  mutate(
    doy = as.numeric(format(as.Date(transition_25),"%j")),
    year = as.numeric(format(as.Date(transition_25),"%Y"))
  ) |>
  select(
    year,
    doy,
    transition_25,
    threshold_25
  )
```

```{r }
# now: Plot the temp. series of all days (distinguish between days where temperature
# threshold is exceeded or not)
# grab only the 2010 value of spring phenology
harvard_phenology_2010 <- harvard_phenology |>
  dplyr::filter(
    year == 2010
  )

harvard_gdd_value <- harvard_temp |>
  dplyr::filter(
    date == harvard_phenology_2010$transition_25
  )

p <- ggplot(harvard_temp) +
  geom_line(
    aes(
      date,
      tmean
    )
  ) +
  geom_point(
    aes(
      date,
      tmean,
      colour = tmean > 5,
      group = 1
    )
  ) +
  geom_vline(
    data = harvard_phenology_2010,
    aes(
      xintercept = as.Date(transition_25)
    )
  ) +
  scale_colour_discrete(
    type = c(
      "blue",
      "red"
    )
  ) +
  labs(
    x = "",
    y = "Temperature (deg. C)"
  ) +
  xlim(
    c(
      as.Date("2010-01-01"),
      as.Date("2010-06-30")
    )
  ) +
  theme_bw() +
  theme(
    legend.position = "none"
  )
```

```{r }
p2 <- ggplot(harvard_temp) +
  geom_line(
    aes(
      date,
      gdd
    )
  ) +
  geom_point(
    aes(
      date,
      gdd,
      colour = tmean > 5,
      group = 1
    )
  ) +
  scale_colour_discrete(
    type = c(
      "blue",
      "red"
    )
  ) +
  geom_vline(
    data = harvard_phenology_2010,
    aes(
      xintercept = as.Date(transition_25)
    )
  ) +
  geom_hline(
    data = harvard_gdd_value,
    aes(
      yintercept = gdd
    ),
    lty = 2
  ) +
  labs(
    x = "",
    y = "GDD (deg. C)"
  ) +
  xlim(
    c(
      as.Date("2010-01-01"),
      as.Date("2010-06-30")
    )
  ) +
  ylim(c(0, 1000)) +
  theme_bw()  +
  theme(
    legend.position = "none"
  )
```

```{r }
# compositing
p + p2 + 
  plot_layout(ncol = 1) + 
  plot_annotation(
    tag_levels = "a",
    tag_prefix = "(",
    tag_suffix = ")"
  )
```

Growing degree day model optimization
```{r }
# function: temp series as first argument, two parameters  (temp threshold above which temperatures are accumulated & critical GDD that determines the DOY at which leaf-out is predicted)
gdd_model <- function(temp, par) {
  # split out parameters from a simple
  # vector of parameter values
  temp_threshold <- par[1]
  gdd_crit <- par[2]
  
  # accumulate growing degree days for
  # temperature data
  gdd <- cumsum(ifelse(temp > temp_threshold, temp - temp_threshold, 0))
  
  # figure out when the number of growing
  # degree days exceeds the minimum value
  # required for leaf development, only
  # return the first value
  doy <- unlist(which(gdd >= gdd_crit)[1])
  
  return(doy)
}
```

```{r }
# confirm that the model function
# returns expected results (i.e. DOY 114)
# (we filter out the year 2010, but
# removing the filter would run the
# model for all years!)
prediction <- harvard_temp |>
  dplyr::filter(
    year == 2010
  ) |>
  group_by(year) |>
  summarize(
    pred = gdd_model(
      temp = tmean,
      par = c(5, 130.44)
    )  
  )
```

```{r }
print(prediction) # a tibble showing the predicted GDD at which the leaf-out was observed

```

## phenology model calibration
```{r }
# goal: minimize the error (cost function) between model results (for 
# given set of parameters (temp & critical GDD) and our observed data 
# estimating parameters efficiently 
## nls()function (square error minimalization for any (non-linear) function)
## or simulated annealing GenSA package
## or Bayesian Tools package
# run model and compare to true values
# returns the RMSE (between observed and predicted values)
rmse_gdd <- function(par, data) {
  
  # split out data
  drivers <- data$drivers
  validation <- data$validation
  
  # calculate phenology predictions
  # and put in a data frame
  predictions <- drivers |>
    group_by(year) |>
    summarise(
      predictions = gdd_model(
        temp = tmean,
        par = par
      )
    )
  
  predictions <- left_join(predictions, validation, by = "year")
  
  rmse <- predictions |>
    summarise(
      rmse = sqrt(mean((predictions - doy)^2, na.rm = TRUE))
    ) |>
    pull(rmse)
  
  # return rmse value
  return(rmse)
}
## !! Both the cost function as the growing degree day function are not optimized for computational efficiency, when implementing your own code, take care to optimize both functions for efficiency by vectorization and other techniques.
```

```{r }
# often, starting model parameters and limits to the parameter space are required (temp. threshold and physiological limits of plant activity)
# starting model parameters
par = c(0, 130)
```

```{r }
# limits to the parameter space
lower <- c(-10,0)
upper <- c(45,500)
```

```{r }
# data needs to be provided in a consistent
# single data file, a nested data structure
# will therefore accept non standard data formats
data <- list(
  drivers = harvard_temp,
  validation = harvard_phenology
)
```

### optimize the model parameters
```{r }
optim_par = GenSA::GenSA(
  par = par,
  fn = rmse_gdd,
  lower = lower,
  upper = upper,
  control = list(
    max.call = 4000 # calling cost function 4000 times
  ),
  data = data
)$par
```

### run the model for all years
plug the optimized parameters back into the model
```{r }
# to get the phenology predictions
predictions <- harvard_temp |>
  group_by(year) |>
  summarize(
    prediction = gdd_model(
      temp = tmean,
      par = optim_par
    )  
  )
```

### results visually
```{r }
# join predicted with observed data
validation <- left_join(predictions, harvard_phenology)

ggplot(validation) +
  geom_smooth(
    aes(
      doy,
      prediction
    ),
    colour = "grey25",
    method = "lm"
  ) +
  geom_point(
    aes(
      doy,
      prediction
    )
  ) +
  geom_abline(
    intercept=0, 
    slope=1, 
    linetype="dotted"
  ) +
  labs(
    x = "Observed leaf-out date (DOY)",
    y = "Predicted leaf-out date (DOY)"
  ) +
  theme_bw()  +
  theme(
    legend.position = "none"
  )
# more advanced models exist: include also radiation, precip, temporal lags (frost during winter)

```

## spatial scaling
```{r, eval=FALSE }
# use model across larger landscape
# 1. Passwort einmal speichern (erstes Mal fragt nach Keyring-Passwort)
key_set(service = "earth_data_user")  
# -> Gib hier dein NASA Earth Data Passwort ein

my_password <- key_get(service = "earth_data_user")
my_password  # zeigt das gespeicherte Passwort an

# appeears_login(user = "earth_data_user", password = my_password)
# if the package has no access to your system password manager, 
# potentially use options(keyring_backend="file") before 
# attempting to set a key
```

```{r, eval=FALSE }
# To see whether the setup has worked, you can try to get the 
# metadata of the product DAYMET v4:
rs_layers("DAYMET.004")
```

```{r, eval=FALSE }
rs_set_key(user = "saja_0610", # your username NASA
           password = "Oleleo2519" # your passwort NASA
)
```

```{r, eval=FALSE }
# Define the bounding box coordinates
xmin <- -72; xmax <- -70
ymin <- 42; ymax <- 44
coords <- matrix(c(
  xmin, ymin,
  xmax, ymin,
  xmax, ymax,
  xmin, ymax,
  xmin, ymin  # Closing the polygon by repeating first point
), ncol = 2, byrow = TRUE)
```

```{r, eval=FALSE }
# Create a simple feature geometry collection with CRS
roi <- st_sf(geometry = st_sfc(st_polygon(list(coords)), 
                               crs = 4326)  # WGS 84)
)
```

```{r, eval=FALSE }
# build the area based request/task
df <- data.frame(
  task = "DAYMET.004_2012",
  subtask = "subtask",
  latitude = NA,
  longitude = NA,
  start = "2012-01-01",
  end = "2012-12-31",
  product = "DAYMET.004",
  layer = c("tmax", "tmin")
)
```

```{r, eval=FALSE }
task <- rs_build_task(
  df = df,
  roi = roi,
  format = "netcdf4"
)
```

request the task to be executed (takes some time -> see https://appeears.earthdatacloud.nasa.gov/explore)
```{r, eval=FALSE }
# request the task to be executed
rs_request(
  request = task,
  user = "saja_0610",
  transfer = TRUE,
  path = here::here("data_raw"),
  verbose = TRUE
)
```

```{r }
# Load the downloaded daily data:
r1 <- terra::rast(
  here::here("data_raw",
             "DAYMET.004_2012", 
             "DAYMET.004_1km_aid0001.nc"), 
  guessCRS=FALSE)
```

```{r }
# Assign the correct CRS, e.g. WGS84 (EPSG:4326)
terra::crs(r1) <- "epsg:4326"
```

The variables can be accessed with:
r1["tmax"]
r1["tmin"]

#nicht machen!!
```{r, eval=FALSE }
## in case there are issues with creating a NASA user account or the download
local_file <- tempfile(fileext = ".nc")
download.file("https://github.com/fabern/handfull_of_pixels/raw/refs/heads/main/data/DAYMET.004_2012/DAYMET.004_1km_aid0001.nc",
              destfile = local_file)
r1 <- terra::rast(local_file)
```

```{r, eval=FALSE }
# Assign the correct CRS, e.g. WGS84 (EPSG:4326)
terra::crs(r1) <- "epsg:4326"
```
#bis hier

jetzt wieder weiter:
```{r }
# Calculate the daily mean values based on 'tmin' and 'tmax'
mean_layer <- terra::mean(r1["tmax"], 
                          r1["tmin"])
```

```{r }
# fix the variable naming
terra::varnames(mean_layer) <- "tmean"                             
names(mean_layer) <- gsub("tmax","tmean",names(mean_layer))
```

```{r }
# plot: result is raster containing a single variable "tmean" for each day
terra::plot(mean_layer)
```

```{r }
# reduce memory footprint of the calculations
# subset to first 180 days
ma_nh_temp <- terra::subset(
  mean_layer,
  1:180
)
```

```{r }
# apply model to this raster (cube)
predicted_phenology <- terra::app(
  ma_nh_temp,
  fun = gdd_model,
  par = optim_par
)
```

## plot results: interactive map of the spatially scaled optimized growing degree model using DAYMET daily mean temperature data
```{r }
# set te colour scale manually
pal <- colorNumeric(
  "magma",
  terra::values(predicted_phenology),
  na.color = "transparent"
)
```

```{r }
# build the leaflet map
# using ESRI tile servers
# and the loaded demo raster
leaflet() |> 
  addProviderTiles(providers$Esri.WorldImagery, group = "World Imagery") |>
  addProviderTiles(providers$Esri.WorldTopoMap, group = "World Topo") |>
  addRasterImage(
    predicted_phenology,
    colors = pal,
    opacity = 0.8,
    group = "Phenology model results"
  ) |>
  addLayersControl(
    baseGroups = c("World Imagery","World Topo"),
    position = "topleft",
    options = layersControlOptions(collapsed = FALSE),
    overlayGroups = c("Phenology model results")
  ) |>
  addLegend(
    pal = pal,
    values = terra::values(predicted_phenology),
    title = "DOY")
```
